---
title: "Converting coordinates to shapefiles"
author: "Christy Dolph"
date: "2025-09-10"
output: html_document
---

# Convert coordinates to shapefile

This script takes site coordinates (longitude, latitude) and converts them to shapefile.

Example data: Claire Bass chloride monitoring sites

## Set up Workspace

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Clean up workspace and load packages (or install if you don't already have them installed)

#Removes everything in work environment
rm(list=ls())


#Packages you need:
want <- c("here", "tidyverse", "readxl", "sf", "tmap", 
          "nhdplusTools", "oce", "tigris")
          
          need <- want[!(want %in% installed.packages()[,"Package"])]
if (length(need)) install.packages(need)
lapply(want, function(i) require(i, character.only=TRUE))
rm(want, need)


#Set up working directories

#Note: Using a workaround for now to set up directories, ultimately we should have all of the data inside the R project folder 

dir <- list()
root.dir <- getwd() #Set Main Directory one level up

#Data Directory:
data.dir <- here("..", "Compiled_data_clean/") 
data.dir

#GIS directory
gis.dir<-here("..", "GIS_layers")
gis.dir

#Map images directory
maps.dir<-here("..", "Maps")

#Water chem directory
dir.chem <- paste(root.dir, "/Water_chemistry_datasets", sep = "")
dir.chem
```

## Load coordinate data:

locations of Claire chloride sampling sites (shared on July 18, 2025)

```{r}

cl.sites<-read_excel(paste0(dir.chem, '/Copy_of_Claire_Bass_Chloride_Sampling_Sites.xlsx'))

head(cl.sites)
```

## Convert coordinates to spatial locations

```{r}

#Note: this sets the coordinate reference system to NAD83 (code=4269)

#Missing values for coordinates are not allowed, so we need to exclude those sites

cl_spatial<-st_as_sf(cl.sites %>% filter(!is.na(Latitude)), coords=c("Longitude", "Latitude"), crs=4269)

#Quick map check:
plot(cl_spatial$geometry)

```

## Write sites to shapefile

```{r}

st_write(cl_spatial, paste0(gis.dir, '/Claire_Bass_Chloride_Sampling_sites.shp'))

```

### Make a nicer map:

Also including Twin Cities Metro Area and stream flowlines

```{r}

#load Twin Cities couty boundaries using tigris package


counties <-counties("Minnesota")
head(counties)
names(counties)

st_crs(counties) #check coordinate reference system 


#Create list of counties in TC Metro 
Metro.List<-c("Anoka", "Carver", "Dakota", "Hennepin", "Ramsey", "Scott", "Washington")

#Filter county shapefile to TC Metro only 
TC.Metro.counties<-counties %>% 
  filter(NAME %in% Metro.List)



#Add in NHD streams layer using nhdplusTools package:
#First dissolve the boundaries between TC counties to create a single contiguous polygon for the area of interest (AOI)
TC_contig <- st_union(TC.Metro.counties)

#Then use this search area to get flowlines:
NHD_TC<-get_nhdplus(
  AOI = TC_contig)


#Plot everything in a map
tm_shape(TC.Metro.counties)+
  tm_borders()+
tm_shape(NHD_TC)+
  tm_lines(col="steelblue1")+
tm_shape(cl_spatial)+
  tm_dots()
```
