---
title: "EPA NARS data processing"
author: "Christy Dolph"
date: "2025-09-16"
output: html_document
---

# Overview: EPA National Aquatic Resource Survey (NARS) data

The following script processes taxonomic data, MMI scores, and accompanying water chem, habitat and spatial data available from the US EPA NARS database for the state of Minnesota. The goal is to be able to analyze trends and drivers of stream biodiversity over space and time, as part of an LCCMR 2024 funded project, 'Impact of statewide conservation practices on biodiversity'.

This data was downloaded directly from EPA at <https://www.epa.gov/national-aquatic-resource-surveys/data-national-aquatic-resource-surveys>

All files for all Rivers and Streams Surveys were downloaded and put in the data/NARS folder of this R project. This corresponds to data from the following NARS survey years: 2004, 2008-2009,2013-2014, 2018-2019.

NARS data last accessed from the web: 2025-09-15

***Tables to process (To do/methods/decisions):***

-   Start with the following data types: taxonomic count, MMI, waterchem

    -   Will need to keep columns that match across datasets

    -   Need to determine whether you need to subsample to 300, or if this was already done across all datasets

    -   Will need to calculate *relative* abundance (not total)

    -   Will need to match taxonomic units across datasets (for species richness)

    -   keep fish and invert data separate for now

    -   do i want to keep MMI condition classes? (see metadata text files)

    -   site tables have coordinates in various formats - eg, 'LON_DD' vs 'LOND83'. the latter makes sense as crs=NAD83. if not listed i assume coordinates in crs=WGS84, and i can convert to NAD83.

    -   note chem files (extensive lab workup) vs DO in 'fieldchem' files

-   Later add trait info - some years (2008) seem to have trait info in separate table, some trait info in MMI tables as well

-   Also see Twardochleb et al database

# Workspace Setup

```{r setup, include=FALSE}

# Clean up workspace 
rm(list=ls())

#install and load packages

#list of necessary packages 
want <- c('dataRetrieval', 'tidyverse', 'sf', 'tmap', 'tigris', 'purrr')


need <- want[!(want %in% installed.packages()[,"Package"])]
if (length(need)) install.packages(need)
lapply(want, function(i) require(i, character.only=TRUE))
rm(want, need)


#Install StreamCatTools (if not already installed):
# First install remotes if needed
if (!requireNamespace("remotes", quietly = TRUE)) {
  install.packages("remotes")
}
# Install StreamCatTools
if (!requireNamespace("StreamCatTools", quietly = TRUE)) {
  remotes::install_github("USEPA/StreamCatTools")
}
# Load the package
library(StreamCatTools)



# Create subdirectory shortcuts

#Main directory
dir.root<-getwd()

#Raw data
dir.raw<-paste0(dir.root, "/Data/biodata_raw")

#NARS data
dir.nars<-paste0(dir.raw, "/NARS")

#Processed data
dir.pro<-paste0(dir.root, "/Data/biodata_processed")
```

# Load NARS data

Select states you would like data processed for:

```{r}
States.list<-c("MINNESOTA")
```

## 2004 Wadeable Streams Assessment (WSA)

Invertebrate data only, no fish. Specific tables:

"wsa_bencnt_genus_ts_final_part1" - taxonomic count data

"wsa_bencnt_genus_ts_final_part2" - taxonomic count data

(what isdifference between part 1 and part 2? they don't contain info for separate sites. maybe just additional count data? can they be combined? check report)

"wsa_siteinfo_ts_final": can be used to match lat/long and sample date by siteID

"wsa_benmet300_ts_final": this is metric info by site (not raw taxonomic data). contains MMI scores.

-Note 'MMI_WSA' is defined as 'CONSISTENT 6 METRIC NATIONWIDE IBI'

\-'MMI_WSABEST': From the 2008 NRSA technical report: "The same autecology and taxonomic resolution used in WSA was applied to the NRSA macroinvertebrate 300 count data to calaculate the community metrics used to calculate the MMI. In the WSA, a best ecoregional MMI was developed by summing the six metrics that performed best in that ecoregion (the national aggregate nine ecoregions). Each of the six metrics was scored on a 0-10 scale by interpolating metrics between a floor and ceiling value. The six metric 0-10 point scaled scores were then summed and normalized to a 0-100 scale by multiplying by 100/60 to calculate the final MMI. Details of this process are described in Stoddard et al. (2008)" <https://www.epa.gov/system/files/documents/2024-10/nrsa-2008-09-final-tsd-version-1.1-september-2024.pdf>

"waterchemistry": extensive water chem analysis

.csv files contain data, .txt files contain metadata on column headers

**Read in taxonomic and IBI data:**

```{r}

# List all files in subfolders that contain "wsa_bencnt"
count.files.2004 <- list.files(
  path = dir.nars,
  pattern = "wsa_bencnt.*\\.csv$",,
  recursive = TRUE,
  full.names = TRUE
)

# Check what was found
print(count.files.2004)

# Read and combine all the files into one data frame
# Assuming they are CSV files — adjust read function if needed
combined_count.2004 <- count.files.2004 %>%
  lapply(read_csv) %>%
  bind_rows() %>% 
  mutate(Assemblage="Benthic macroinvertebrates")#create assemblage column

# View the combined data
glimpse(combined_count.2004)

#View(combined_count.2004)

#####################################################
#Load in site info
sites2004<-read.csv(list.files(
  path = dir.nars,
  pattern = "wsa_siteinfo.*\\.csv$",,
  recursive = TRUE,
  full.names = TRUE
))
head(sites2004) 

#NOTE: convert coordinates to crs=NAD83
#assume coordinates initially taken in WGS84
sites2004_sf<-st_as_sf(sites2004, coords=c('LON_DD', 'LAT_DD'), crs=4326)

#transform to NAD83
sites2004.NAD83<-st_transform(sites2004_sf, crs=4269)

#Quick map check
plot(sites2004.NAD83$geometry)

#Add coordinates in NAD83 as separate columns
nars_sites2004<-sites2004.NAD83 %>%
  dplyr::mutate(LON_DD83 = sf::st_coordinates(.)[,1],
                LAT_DD83 = sf::st_coordinates(.)[,2]) %>% 
  st_drop_geometry()

#View(nars_invert_2004)


###################################################### 
#Merge site info to taxonomic data for select state (ie, MN)
names(combined_count.2004)
names(sites2004)

site.invert.2004<-merge(
  nars_sites2004 %>% 
    dplyr::select(SITE_ID, SITENAME, SITETYPE, REPEAT, LON_DD83, LAT_DD83, XCOORD_SOURCE, STATE_NM, DATE_COL, STRAHLER, WSAREA, FLOWSITE) %>% 
    filter(STATE_NM %in% States.list), combined_count.2004, by=c("SITE_ID"))

#View(site.invert.2004)


#Check count data across sites
names(site.invert.2004)
site.invert.2004 %>% group_by(SITE_ID) %>% summarise(total=sum(ABUND))
#Note these are not subampled to 300

####################################################
#Read in MMI data:
MMI.2004<-read.csv(list.files(
  path = dir.nars,
  pattern = "benmet.*\\.csv$",,
  recursive = TRUE,
  full.names = TRUE
))
head(MMI.2004) 
#View(MMI.2004)
names(MMI.2004)

MMI.2004<-MMI.2004 %>% 
  mutate(Assemblage="Benthic macroinvertebrates")

#Select desired metrics and desired region:

site.MMI.invert.2004<-merge(nars_sites2004 %>% 
    dplyr::select(SITE_ID, SITENAME, SITETYPE, REPEAT, LON_DD83, LAT_DD83, XCOORD_SOURCE, STATE_NM, DATE_COL, STRAHLER, WSAREA, FLOWSITE) %>% 
    filter(STATE_NM %in% States.list),
    
    MMI.2004 %>% dplyr::select(SITE_ID, MMI_WSA, MMI_WSABEST, Assemblage), by=c("SITE_ID")) 

#View(site.MMI.invert.2004)

```

## 2008-2009 National Rivers and Streams Assessment

Files listed as "external data", start with "ext" in file name - what is the difference with the "regular" data? - External data appears to contain data from previous years (back to 1996 for MN, includes 2004 WSA data), but not the current sample year

MMI for fish are complicated -

-They made an RF model to predict fish MMI based on least disturbed sites. "The predictive model for each metric was then applied to the entire set of sites, and the residuals (deviation from predicted) were used as the response value for the metric." (from 2008 NRSA technical report). So they have columns for each residual for each metric in the relevant MMI.

\-"FISH_MMI_RF_COND" = fish assemblage condition, based on regional random forest Fish MMI

-Skipping Fish MMI compilation for now because of this complexity

```{r}

#Read in taxonomic data:

invert.count.2008<-
  read.csv(list.files(
  path = dir.nars,
  pattern = "nrsa0809bentcts.csv",,
  recursive = TRUE,
  full.names = TRUE
)) %>% mutate(Assemblage="Benthic macroinvertebrates")

fish.count.2008<-read.csv(list.files(
  path = dir.nars,
  pattern = "fishcts.csv",,
  recursive = TRUE,
  full.names = TRUE
)) %>%  mutate(Assemblage="Fish")

names(invert.count.2008)
head(invert.count.2008)
names(fish.count.2008)

###############################################

#Read in Site info:
site.2008<-
  read.csv(list.files(
  path = dir.nars,
  pattern = "siteinfo_0.csv",,
  recursive = TRUE,
  full.names = TRUE
))
names(site.2008)
head(site.2008)

#####################################################
#Merge site to taxonomic info 
site.invert.2008<-merge(site.2008 %>% dplyr::select(SITE_ID, STATE, LOC_NAME, SITE_CLASS, LAT_DD83, LON_DD83, DATE_COL, STRAHLERORDER, WSAREA_NARS ) %>% filter(STATE=="MN"), invert.count.2008, by=c("SITE_ID", "DATE_COL"))
#View(site.invert.2008)

site.fish.2008<-merge(site.2008 %>% dplyr::select(SITE_ID, STATE, LOC_NAME, SITE_CLASS, LAT_DD83, LON_DD83, DATE_COL, STRAHLERORDER, WSAREA_NARS) %>% filter(STATE=="MN"), fish.count.2008, by=c("SITE_ID", "DATE_COL"))


#####################################################

#MMI scores:
MMI.invert.2008<-
  read.csv(list.files(
  path = dir.nars,
  pattern = "bentcond.csv",,
  recursive = TRUE,
  full.names = TRUE
)) %>% dplyr::select(SITE_ID, VISIT_NO, DATE_COL, MMI_BENT) 
  
#Merge to site info
site.MMI.invert.2008<-merge(site.2008 %>% dplyr::select(SITE_ID, STATE, LOC_NAME, SITE_CLASS, LAT_DD83, LON_DD83, DATE_COL, STRAHLERORDER, WSAREA_NARS) %>% filter(STATE=="MN"), 
MMI.invert.2008, by=c("SITE_ID", "DATE_COL"))

```

## 2013-2014 National Rivers and Streams Assessment

Notes:

-they did subsample for 300 individuals in a separate column, but may want to keep total count and then subsample separately to be consistent with other years? tho note their MMI might be based on the 300 subcount?

-they also include an O/E score

-note very extensive chemistry in 'widechem' file

```{r}

#Load taxonomic data
invert.count.2013<-
  read.csv(list.files(
  path = dir.nars,
  pattern = "nrsa1314_bentcnts.*\\.csv$",,
  recursive = TRUE,
  full.names = TRUE
)) %>% mutate(Assemblage="Benthic macroinvertebrates")

fish.count.2013<-read.csv(list.files(
  path = dir.nars,
  pattern = "nrsa1314_fishcts.*\\.csv$",,
  recursive = TRUE,
  full.names = TRUE
)) %>%  mutate(Assemblage="Fish")

names(invert.count.2013)
names(fish.count.2013)

###########################################
#Load site info:
site.2013<-
  read.csv(list.files(
  path = dir.nars,
  pattern = "nrsa1314_site.*\\.csv$",,
  recursive = TRUE,
  full.names = TRUE
))
names(site.2013)

##########################################
#Merge taxonomic and site info

site.invert.2013<-merge(
  site.2013 %>% dplyr::select(SITE_ID, DATE_COL, VISIT_NO, STATE, LAT_DD83, LON_DD83, STRAH_ORD, COMID) %>% filter(STATE=="MN"), 

  invert.count.2013, by=c("SITE_ID", "VISIT_NO"))


site.fish.2013<-merge(
  site.2013 %>% dplyr::select(SITE_ID, DATE_COL, VISIT_NO, STATE, STRAH_ORD, COMID) %>% filter(STATE=="MN"), 
  
  fish.count.2013, by=c("SITE_ID", "VISIT_NO", "DATE_COL"))

####################################

#MMI scores:
MMI.invert.2013<-
  read.csv(list.files(
  path = dir.nars,
  pattern = "nrsa1314_bentmmi.*\\.csv",,
  recursive = TRUE,
  full.names = TRUE
)) %>% dplyr::select(SITE_ID, VISIT_NO, DATE_COL, MMI_BENT) %>% mutate(Assemblage="Benthic macroinvertebrates")

MMI.fish.2013<-read.csv(list.files(
  path = dir.nars,
  pattern = "nrsa-1819-fish-mmi-data.csv",,
  recursive = TRUE,
  full.names = TRUE
)) %>% dplyr::select(SITE_ID, VISIT_NO, DATE_COL, MMI_FISH)%>% mutate(Assemblage="Fish") 
  
#Merge to site info
site.MMI.invert.2013<-merge(site.2013 %>% dplyr::select(SITE_ID, STATE, LAT_DD83, LON_DD83, DATE_COL, COMID) %>% filter(STATE=="MN"), 
MMI.invert.2013, by=c("SITE_ID", "DATE_COL"))

site.MMI.fish.2013<-merge(site.2013 %>% dplyr::select(SITE_ID, STATE, LAT_DD83, LON_DD83, DATE_COL, COMID) %>% filter(STATE=="MN"), 
MMI.fish.2013, by=c("SITE_ID", "DATE_COL"))

```

## 2018 National Rivers and Streams Assessment

```{r}
#Load taxonomic data
invert.count.2018<-
  read.csv(list.files(
  path = dir.nars,
  pattern = "nrsa_1819_benthic_macroinvertebrate_count_-_data.*\\.csv$",,
  recursive = TRUE,
  full.names = TRUE
)) %>% mutate(Assemblage="Benthic macroinvertebrates")

fish.count.2018<-read.csv(paste0(dir.nars, "/NARS_NRSA_2018_2019/nrsa-1819-fish-count-data.csv"))%>% mutate(Assemblage="Fish")

names(invert.count.2018)
names(fish.count.2018)

###########################################
#Load site info:
site.2018<-
  read.csv(list.files(
  path = dir.nars,
  pattern = "NRSA_1819_SiteInfo.*\\.csv$",,
  recursive = TRUE,
  full.names = TRUE
))
names(site.2018)

site.2018 %>% dplyr::select(STATE_SF, STATE_NM, SF_ST_ID) %>% head()

##########################################
#Merge taxonomic and site info

site.invert.2018<-merge(
  site.2018 %>% dplyr::select(SITE_ID, DATE_COL, VISIT_NO, STATE_NM, LAT_DD83, LON_DD83, STRAH_ORD, COMID) %>% filter(STATE_NM=="Minnesota"), 
  invert.count.2018, by=c("SITE_ID", "DATE_COL"))


site.fish.2018<-merge(
  site.2018 %>% dplyr::select(SITE_ID, DATE_COL, VISIT_NO, STATE_NM, LAT_DD83, LON_DD83, STRAH_ORD, COMID) %>% filter(STATE_NM=="Minnesota"), 
  
  fish.count.2018, by=c("SITE_ID", "DATE_COL"))

####################################

#MMI scores:
#MMI scores:
MMI.invert.2018<-
  read.csv(list.files(
  path = dir.nars,
  pattern = "nrsa-1819-benthic-macroinvertebrate-mmi-data.csv",,
  recursive = TRUE,
  full.names = TRUE
)) %>% dplyr::select(SITE_ID, VISIT_NO, DATE_COL, MMI_BENT) %>% mutate(Assemblage="Benthic macroinvertebrates")

MMI.fish.2018<-read.csv(list.files(
  path = dir.nars,
  pattern = "nrsa-1819-fish-mmi-data.csv",,
  recursive = TRUE,
  full.names = TRUE
)) %>% dplyr::select(SITE_ID, VISIT_NO, DATE_COL, MMI_FISH)%>% mutate(Assemblage="Fish") 
  
#Merge to site info
site.MMI.invert.2018<-merge(site.2018 %>% dplyr::select(SITE_ID, STATE_NM, LAT_DD83, LON_DD83, DATE_COL, COMID) %>% filter(STATE_NM=="Minnesota"), 
MMI.invert.2018, by=c("SITE_ID", "DATE_COL"))

site.MMI.fish.2018<-merge(site.2018 %>% dplyr::select(SITE_ID, STATE_NM, LAT_DD83, LON_DD83, DATE_COL, COMID) %>% filter(STATE_NM=="Minnesota"), 
MMI.fish.2018, by=c("SITE_ID", "DATE_COL"))

```

# Combine data across NARS survey years

## Benthic Macroinvertbrates

```{r}

#relevant tables - 
tax.tables<-c(site.invert.2004,
site.invert.2008,
site.invert.2013,
site.invert.2018)

#reformat names and columns as needed for each table

names(site.invert.2004)
names(site.invert.2008)

site.invert.2004_CLEAN<-site.invert.2004 %>% 
  mutate(Date=as.Date(DATE_COL, format="%m/%d/%Y")) %>%  #reformat Date
  rename(TOTAL=ABUND) %>% #rename total column 
  mutate(TOTAL300=NA) %>% #create 300 subsample column
  mutate(TARGET_TAXON=NA) %>% #create empty target taxon column %>% 
  mutate(SUBFAMILY=NA, TRIBE=NA) %>% #make empty taxonomic columns
  mutate(TARGET_TAXON=coalesce(GENUS, FAMILY, ORDER, CLASS, PHYLUM)) %>%   #Assign TARGET_TAXON with most specific ID available; select in this order: GENUS, FAMILY, ORDER, CLASS PHYLUM
  dplyr::select(SITE_ID, Date, LON_DD83, LAT_DD83, TARGET_TAXON, TOTAL, TOTAL300, Assemblage, PHYLUM, CLASS, ORDER, FAMILY, SUBFAMILY, TRIBE, GENUS) #select desired columns
  
  
#View(site.invert.2008)    
site.invert.2008_CLEAN<-site.invert.2008 %>% 
  mutate(Date=as.Date(DATE_COL, format="%d-%b-%y")) %>%  
  mutate(Date=as.Date(Date, format="%m-%d-%Y")) %>%  
  mutate(PHYLUM=NA, CLASS=NA, ORDER=NA, FAMILY=NA, SUBFAMILY=NA, TRIBE=NA, GENUS=NA) %>% #make empty taxonomic columns
dplyr::select(SITE_ID, Date, LON_DD83, LAT_DD83, TARGET_TAXON, TOTAL, TOTAL300, Assemblage, PHYLUM, CLASS, ORDER, FAMILY, SUBFAMILY, TRIBE,  GENUS)  #select desired columns
  

#View(site.invert.2008)

names(site.invert.2013)
site.invert.2013_CLEAN<-site.invert.2013 %>% 
   mutate(Date=as.Date(DATE_COL, format="%m/%d/%Y")) %>% #reformat Date
  dplyr::select(SITE_ID, Date, LON_DD83, LAT_DD83, TARGET_TAXON, TOTAL, TOTAL300, Assemblage, PHYLUM, CLASS, ORDER, FAMILY, SUBFAMILY, TRIBE, GENUS)

names(site.invert.2018)

site.invert.2018_CLEAN<-site.invert.2018 %>% 
   mutate(Date=as.Date(DATE_COL, format="%m/%d/%Y")) %>% #reformat Date
  dplyr::select(SITE_ID, Date, LON_DD83, LAT_DD83, TARGET_TAXON, TOTAL, TOTAL300, Assemblage, PHYLUM, CLASS, ORDER, FAMILY, SUBFAMILY, TRIBE, GENUS)

#Combine all data! 

nars.inverts<-rbind(site.invert.2004_CLEAN, site.invert.2008_CLEAN, site.invert.2013_CLEAN, site.invert.2018_CLEAN)

#View(nars.inverts)

#Write to file

write.csv(nars.inverts, paste0(dir.pro, "/nars_invert_data_all.csv"))

#Count number of sites by taxon and year
#count sites
nars.inverts %>% 
  mutate(Year=format(Date, "%Y")) %>% 
  dplyr::select(SITE_ID, Year) %>% 
  distinct() %>% group_by(Year) %>% count()



```

## Fish

```{r}

#relevant tables
fish.tables<-c(site.fish.2008,
site.fish.2013,
site.fish.2018)

names(site.fish.2008)
names(site.fish.2013)
names(site.fish.2018)

#View(site.fish.2013)

#Format columns across datasets:
#Need to format Date (different formatting for different years)
site.fish.2008_CLEAN<-site.fish.2008 %>% 
  mutate(Date=as.Date(DATE_COL, format="%d-%b-%y")) %>%  
  mutate(Date=as.Date(Date, format="%m-%d-%Y")) %>% 
  mutate(TOTAL=FINAL_CT) #format fish count

site.fish.2013_CLEAN<-site.fish.2013 %>% 
  mutate(Date=as.Date(DATE_COL, format="%m/%d/%Y"))

site.fish.2018_CLEAN<-site.fish.2018 %>% 
  mutate(Date=as.Date(DATE_COL, format="%m/%d/%Y"))
 
#column names are the same across years;
#select desired columns

col.want<-c("SITE_ID", "Date", "LAT_DD83", "LON_DD83", "TAXA_ID", "FINAL_NAME", "TOTAL", "Assemblage")

nars.fish <- bind_rows(
site.fish.2008_CLEAN %>% 
  dplyr::select(all_of(col.want)), 
site.fish.2013_CLEAN %>% 
  dplyr::select(all_of(col.want)),
site.fish.2018_CLEAN %>% 
  dplyr::select(all_of(col.want))
)

View(nars.fish)

#Write to file

write.csv(nars.fish, paste0(dir.pro, "/nars_fish_data_all.csv"))


#Count number of sites by taxon and year
#count sites
nars.fish %>% 
  mutate(Year=format(Date, "%Y")) %>% 
  dplyr::select(SITE_ID, Year) %>% 
  distinct() %>% group_by(Year) %>% count()


```

## Chemistry

-   All NARS survey years have a 'water chem' file (water samples analyzed in a lab) and a 'field chem' file (parameters sampled in the field including DO, pH, temp)

-   Note metadata files on NARS website for column definitions and unites:

<https://www.epa.gov/national-aquatic-resource-surveys/data-national-aquatic-resource-surveys>

-   *Note change in units between 2004-2008*

-   Note that 2004 NO3 just listed as 'nitrate', 2008 and 2013 listed as 'NO3 as nitrogen'

-   Note the many alerts on many water chem parameters - exceeding holding time, shipping time

-   some variability in water parameter availability across years (eg, HCO3 in 2004 but not 2008, NO3NO2 in 2008 but not 2004)

-   there is anion data available for 2004- not sure if this is useful

-   2004:

    -   PTL = Total P (ug/L), NTL=Total N (ug/L)

    -   PHSTVL = Closed headspace ph

    -   note some 2004 units are mg/L and some are ug/L

### Read in water chemistry files

```{r}
####################################################
#Read in waterchem data:
chem.2004<-read.csv(list.files(
  path = dir.nars,
  pattern = "waterchemistry.*\\.csv$",,
  recursive = TRUE,
  full.names = TRUE
))
names(chem.2004)

field.chem.2004<-read.csv(paste0(dir.nars, "/NARS_2004_Wadeable_Stream_Assessment/fieldchemistry.csv"))

##################################
#Water Chem: 
chem.2008<-read.csv(paste0(dir.nars, "/NARS_NRSA_2008_2009/chem.csv"))
names(chem.2008)

field.chem.2008<-read.csv(paste0(dir.nars, "/NARS_NRSA_2008_2009/fieldchemmeasure.csv"))
 
#######################################
#Water Chem: 
chem.2013<- read.csv(list.files(
  path = dir.nars,
  pattern = "nrsa1314_widechem.*\\.csv$",,
  recursive = TRUE,
  full.names = TRUE
))

field.chem.2013<-read.csv(list.files(
  path = dir.nars,
  pattern = "nrsa1314_wide_field_meas.*\\.csv$",,
  recursive = TRUE,
  full.names = TRUE
))

#######################################
#Water Chem: 
chem.2018<- read.csv(list.files(
  path = dir.nars,
  pattern = "nrsa_1819_water_chemistry_chla_-_data.*\\.csv$",,
  recursive = TRUE,
  full.names = TRUE
))

field.chem.2018<-read.csv(list.files(
  path = dir.nars,
  pattern = "NRSA_1819_field_wide.*\\.csv$",,
  recursive = TRUE,
  full.names = TRUE
))
#View(field.chem.2018)
```

### Main water chemistry (samples analyzed in lab)

```{r}
chem.tables<-c(chem.2004, chem.2008, chem.2013, chem.2018)

names(chem.2004)


#Select common columns and reformat names:
#convert units to from ueq/L to mg/L where needed using conversion factors 
#conversion factors are based on mg/L=(ueq/L x Equivalent weight)/1000
#Equivalent weight=Molecular weight/valence
#convert from ug/L to mg/L
#convert NO3 and NH4 to NO3-N and NH4-N to be consistent with other NARS years 

chem.2004.clean<-chem.2004 %>% 
mutate(
  COND_uS_cm = COND, 
  CA_mg_L = CA * 20.04/1000, 
  CL_mg_L = CL * 35.45/1000, 
  TSS_mg_L = TSS,
  TURB_NTU = TURB,
  DOC_mg_L = DOC,
  #DIC_mg_L = DIC,
  #HCO3_mg_L = HCO3 * 61.02,
  K_mg_L = K * 39.1/1000,
  MG_mg_L = MG * 12.15/1000,
  NA_mg_L = NA. * 22.99/1000,
  NH4_mg_L = NH4 * 18.04*0.7766/1000, #convert to NH4-N
  NO3N_mg_L = NO3 * 62*0.226/1000, #also convert to NO3-N
  PH=PHSTVLF,
  TN_mg_L = NTL / 1000,
  TP_mg_L=PTL  / 1000,
  #ZN_mg_L=ZN * 1000,
  SO4_mg_L=SO4 * 48.03/1000,
  SIO2_mg_L=SIO2) %>% #rename for units 
dplyr::select(SITE_ID, DATE_COL, COND_uS_cm, 
  CA_mg_L, CL_mg_L, TSS_mg_L, TURB_NTU, DOC_mg_L, K_mg_L, MG_mg_L, NA_mg_L, NH4_mg_L, NO3N_mg_L, PH, TN_mg_L, TP_mg_L, SO4_mg_L, SIO2_mg_L)%>% #select chem variables
  mutate(across(everything(as.numeric))) %>% 
  mutate(PH=as.numeric(PH))#convert to numeric 

#gut check after converting
summary(chem.2004.clean$NO3N_mg_L) #note extremely high outlying values
summary(chem.2004.clean$TP_mg_L)  

names(chem.2008)

chem.2008.clean<-chem.2008 %>% 
mutate(
  COND_uS_cm = COND, 
  CA_mg_L = CA, 
  CL_mg_L = CL, 
  TSS_mg_L = TSS,
  TURB_NTU = TURB,
  DOC_mg_L = DOC,
  #DIC_mg_L = DIC,
  #HCO3_ueq_L = HCO3,
  K_mg_L = K,
  MG_mg_L = MG,
  NA_mg_L = SODIUM,
  NH4N_mg_L = NH4,
  NO3N_mg_L = NO3,
  PH=PHLAB,
  TN_mg_L = NTL / 1000,
  TP_mg_L=PTL / 1000,
  #ZN_ug_L=ZN,
  SO4_mg_L=SO4,
  SIO2_mg_L=SIO2) %>%  #rename for units 
dplyr::select(SITE_ID, DATE_COL, COND_uS_cm, 
  CA_mg_L, CL_mg_L, TSS_mg_L, TURB_NTU, DOC_mg_L, K_mg_L, MG_mg_L, NA_mg_L, NH4N_mg_L, NO3N_mg_L, PH, TN_mg_L, TP_mg_L, SO4_mg_L, SIO2_mg_L)%>% 
  mutate(across(everything(as.numeric))) #convert to numeric

#gut check
summary(chem.2008.clean$NO3N_mg_L)
summary(chem.2008.clean$TP_mg_L)

names(chem.2013)
chem.2013.clean<-chem.2013 %>% 
   mutate(
  COND_uS_cm = COND_RESULT,
  CA_mg_L = CALCIUM_RESULT, 
  CL_mg_L = CHLORIDE_RESULT, 
  TSS_mg_L = TSS_RESULT,
  TURB_NTU = TURB_RESULT,
  DOC_mg_L = DOC_RESULT,
  #DIC_mg_L = DIC,
  #HCO3_ueq_L = HCO3,
  K_mg_L = POTASSIUM_RESULT,
  MG_mg_L = MAGNESIUM_RESULT,
  NA_mg_L = SODIUM_RESULT,
  NH4N_mg_L = AMMONIA_N_RESULT,
  NO3N_mg_L = NITRATE_N_RESULT,
  PH = PH_RESULT,
  TN_mg_L = NTL_RESULT,
  TP_mg_L=PTL_RESULT/1000,
  #ZN_ug_L=ZN,
  SO4_mg_L=SULFATE_RESULT,
  SIO2_mg_L=SILICA_RESULT) %>% #rename for units 
dplyr::select(SITE_ID, DATE_COL, COND_uS_cm, 
  CA_mg_L, CL_mg_L, TSS_mg_L, TURB_NTU, DOC_mg_L, K_mg_L, MG_mg_L, NA_mg_L, NH4N_mg_L, NO3N_mg_L, PH, TN_mg_L, TP_mg_L, SO4_mg_L, SIO2_mg_L)%>% 
  mutate(across(everything(as.numeric))) #convert to numeric

summary(chem.2013.clean$TP_mg_L)

names(chem.2018)

chem.2018.clean<-chem.2018 %>% 
  mutate(
  COND_uS_cm = COND_RESULT,
  CA_mg_L = CALCIUM_RESULT, 
  CL_mg_L = CHLORIDE_RESULT, 
  TSS_mg_L = TSS_RESULT,
  TURB_NTU = TURB_RESULT,
  DOC_mg_L = DOC_RESULT,
  #DIC_mg_L = DIC,
  #HCO3_ueq_L = HCO3,
  K_mg_L = POTASSIUM_RESULT,
  MG_mg_L = MAGNESIUM_RESULT,
  NA_mg_L = SODIUM_RESULT,
  NH4N_mg_L = AMMONIA_N_RESULT,
  NO3N_mg_L = NITRATE_N_RESULT,
  PH=PH_RESULT,
  TN_mg_L = NTL_RESULT,
  TP_mg_L=PTL_RESULT/1000,
  #ZN_ug_L=ZN,
  SO4_mg_L=SULFATE_RESULT,
  SIO2_mg_L=SILICA_RESULT) %>% #rename for units 
dplyr::select(SITE_ID, DATE_COL, COND_uS_cm, 
  CA_mg_L, CL_mg_L, TSS_mg_L, TURB_NTU, DOC_mg_L, K_mg_L, MG_mg_L, NA_mg_L, NH4N_mg_L, NO3N_mg_L, PH, TN_mg_L, TP_mg_L, SO4_mg_L, SIO2_mg_L) %>% 
  mutate(across(everything(as.numeric))) %>% 
  mutate(TSS_mg_L=as.numeric(TSS_mg_L))#convert to numeric

#Combine all chem data
nars.chem<-bind_rows(chem.2004.clean, chem.2008.clean, chem.2013.clean, chem.2018.clean)

View(nars.chem)


```

### Field chemistry

-Includes DO, pH, temp

-Note there are quite a lot of comments for DO in 2004 survey year - meter not working, etc, may want to keep flags;

-Omitting field measured pH as you have lab measured pH from above and not all years have field-measured

-Some values are calibration, and some are field measure - select only field measure

-check DO units (mg/L vs %) - assume if units not listed it is mg/L; 2018 is only data that provides units; exclude % saturation values

```{r}
field.chem.tables<-c(field.chem.2004, field.chem.2008, field.chem.2013, field.chem.2018)

names(field.chem.2004)
names(field.chem.2008)
names(field.chem.2013)
names(field.chem.2018)

levels(factor(field.chem.2013$SAMPLE_TYPE))
levels(factor(field.chem.2018$SAMPLE_TYPE))

nars.field.chem<-bind_rows(
field.chem.2004 %>% rename(DO=STRM_DO, TEMP=STRMTEMP, FLAG=COM_DO) %>% dplyr::select(SITE_ID, DATE_COL, DO, TEMP, FLAG) %>% mutate(across(c(DO, TEMP), as.numeric)),

field.chem.2008 %>% rename(TEMP=TEMPERATURE, FLAG=COMMENT) %>% dplyr::select(SITE_ID, DATE_COL, DO, TEMP, FLAG) %>% mutate(across(c(DO, TEMP), as.numeric)),

field.chem.2013 %>% rename(TEMP=TEMPERATURE, FLAG=DO_FLAG) %>% filter(SAMPLE_TYPE=="FIELDMEAS")%>% dplyr::select(SITE_ID, DATE_COL, DO, TEMP, FLAG)%>% mutate(across(c(DO, TEMP), as.numeric)),

field.chem.2018 %>% rename(TEMP=TEMPERATURE) %>% 
  filter(SAMPLE_TYPE=="FIELDMEAS") %>% 
  filter(DO_DISPLAYED_UNITS=="mg/L") %>% 
  dplyr::select(SITE_ID, DATE_COL, DO, TEMP, FLAG) %>% mutate(across(c(DO, TEMP), as.numeric))
)

View(nars.field.chem)

#Gut check of values
summary(nars.field.chem$DO)

nars.field.chem<-nars.field.chem %>% 
mutate(FLAG = ifelse(FLAG == "", NA, FLAG)) %>% 
mutate(FLAG = ifelse(FLAG == " ", NA, FLAG)) 

#Look only at DO values with flags:
View(nars.field.chem %>% 
       filter(!is.na(FLAG)))

#Omit any row that flag DO issues, ie that contain 'DO' or 'D.O.'

nars.field.chem<-nars.field.chem %>% 
  filter(!grepl("DO|D.O.", FLAG))
```

Merge water chemistry and field chemistry and write to file

```{r}

nars.chem.all<-merge(nars.chem, nars.field.chem, by=c("SITE_ID", "DATE_COL"))

#write to file
write.csv(nars.chem.all, paste0(dir.pro, "/nars_water_chemistry.csv"))

```
