---
title: "Mapping_urban_biodata"
author: "Christy Dolph"
date: "2025-04-24"
output: html_document
editor_options: 
  chunk_output_type: inline
---

# Overview

This script contains methods for spatial mapping of biomonitoring data assembled from local agencies in the Twin Cities Metro Area. It does the following:

-   Option to load a single file or all the clean data files

-   If all data files are loaded, merges them into one dataset. To do this, the script first tries to clean up the column names so they match across datasets. (tho notes as of 4.24 this is still creating problems because columns are named inconsistently across input files)

-   Creates Longitude/Latitude in decimal degrees if these are missing

-   Plots a map of the data

# Workspace setup

Load required packages and create directories

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Clean up workspace and load packages (or install if necessary)
#rm(list=ls())
want <- c("here", "tidyverse", "readxl", "sf", "tmap", 
          "nhdplusTools", "oce", "tigris")
          
          need <- want[!(want %in% installed.packages()[,"Package"])]
if (length(need)) install.packages(need)
lapply(want, function(i) require(i, character.only=TRUE))
rm(want, need)


#Set up working directories

#Note: Using a workaround for now to set up directories, ultimately we should have all of the data inside the R project folder 

dir <- list()
root.dir <- getwd() #Set Main Directory one level up
list.files(root.dir)

#Data Directory:
data.dir <- paste0(root.dir, "/Data/biodata_raw/TC_Metro/Compiled_data_clean") 
data.dir

#GIS directory
gis.dir<-paste0(root.dir, "/Data/GIS")
gis.dir

#Map images directory
maps.dir<-paste0(root.dir, "/Maps")
maps.dir

#Water chem directory
dir.chem <- paste0(root.dir, "/Data/water_chemistry")
dir.chem

#Output data
dir.output<-paste0(root.dir, "/Data/biodata_processed")
dir.output
```

# Pre-processed biomonitoring data:

These files have all already undergone one round of preprocessing/cleaning. This pre-processing included some or all of the following (depending on the state of the original raw data):

-   compiling data from multiple input files (and/or multiple excel tabs) per agency into one file

-   creating common column across all input datasets

-   matching sampling sites to coordinates (ie, spatial locations)

-   removing idiosyncratic file formatting and converting all data into long form (one data entry per row with all relevant information)

## Load in data for all the individual files separately, then merge

Load in pre-processed biomonitoring data from each agency as separate files, then merge.

```{r}

#list all files in the compiled data folder:
list.files(data.dir)

#read in all relevant files (including excel and csv files)
#and ignore temporary files

final_files<-list.files(path=data.dir, pattern = "^[^~].*\\.(csv|xlsx)$", full.names = TRUE, ignore.case = TRUE
)
final_files


#Exclude files that aren't sufficiently processed (i.e., don't have coordinate data etc)

remove_keywords <- c(
  'Blaine Biomonitoring Fall 2023 Data',
  'gsheet'
)

# Keep only those that do NOT partially match any of the keywords
filter_files <- final_files[!sapply(final_files, function(x) (grepl(paste(remove_keywords, collapse = "|"), x)))]
filter_files

# Function to read each file depending on its extension
read_file <- function(file_path) {
  if (grepl("\\.csv$", file_path, ignore.case = TRUE)) {
    read.csv(file_path, header=TRUE)
  } else if (grepl("\\.xlsx$", file_path, ignore.case = TRUE)) {
    read_excel(file_path, col_types="text")
  } else {
    NULL
  }
}



# Read all files into a list (ignoring temporary/open files)
data_list <- lapply(filter_files, read_file)

data_list

#Name each element in the list with the base file name (without extension)
names(data_list) <- tools::file_path_sans_ext(basename(filter_files))

# Standardize and select relevant columns from each file 

# Desired columns
#convert to lowercase column names for matching

target_cols<-c("Original_ID",	"Collection_Date", "Season",	"Year",	"Replicate",	"Agency",	"Collected_by",	"Class", "Subclass", "Order", "Family",	"Genus", "Species", "Original_Taxon",	"Number",	"Notes", "TSN",	"Site", "Latitude", "Longitude", "Northing", "Easting")

target_cols_lower <- tolower(target_cols)
target_cols_lower


# Function to process and standardize each table
process_table <- function(df) {
  # Rename columns to lowercase for easier matching
  names(df) <- tolower(names(df))
  # Add missing columns
  missing_cols <- setdiff(target_cols_lower, names(df))
  for (col in missing_cols) {
    df[[col]] <- NA
  }
  
  # Select and coerce columns
  df %>%
    dplyr::select(all_of(target_cols_lower)) %>%
    mutate(
      original_id = as.character(original_id),
      date = as.Date(round(as.numeric(collection_date),0), origin = "1899-12-30"), #issues with date formatting
      year = as.numeric(year),
      replicate=as.numeric(replicate),
      number = as.numeric(number),
      tsn=as.numeric(tsn),
      longitude=as.numeric(longitude), 
      latitude=as.numeric(latitude),
      northing=as.numeric(northing),
      easting=as.numeric(easting)
    )
}


# Read, process, and combine all tables
combined_data <- filter_files %>%
  map(read_file) %>%
  map(process_table) %>%
  bind_rows()

View(combined_data)


#Rename combined data so that script below will work:
biodata<-combined_data

#Check agencies included
levels(factor(biodata$agency))

#Check for NAs in agency
biodata %>% filter(is.na(agency)) %>% count()

```

### Convert UTM to Long/Lat in decimal degrees:

"Easting" (also sometimes called UTMx) is equivalent to Longitude

"Northing" (also sometimes called UTMy) is equivalent to Latitude

In Minnesota we are in UTM zone 15N

Use utm2lonlat() function to convert to Long/Lat in WGS84

(Note that because this creates coordinates in the coordinate reference system WGS84 we will need to transform to NAD83 to match other data)

For sites that have upstream and downstream coordinates - use the downstream coordinates as the final coordinates

```{r}

#For this to work properly, it would be good if all input files had both lat/long and northing/easting columns (even if they don't have any data)

nrow(biodata)
names(biodata)

biodata


biodata.coord<-biodata %>% 
  mutate_at(vars(longitude, latitude, easting, northing, number), as.numeric) %>% #set specified columns to numeric
  #*IF* long/lat columns are empty, convert longitude and latitude from UTM:
  #create longitude column in decimal degrees:
  mutate(longitude=ifelse(is.na(longitude), (utm2lonlat(easting, northing, zone=15, hemispher="N")$longitude), longitude)) %>% 
  #create Latitute column in decimal degrees:
  mutate(latitude=ifelse(is.na(latitude), (utm2lonlat(easting, northing, zone=15, hemispher="N")$latitude), latitude)) %>%
#*IF* long/lat columns have upstream/downstream designation, choose downstream coordinates as final coordinates:
#  mutate(ifelse(is.na(latitude), down_latitude, latitude)) %>% 
#  mutate(ifelse(is.na(longitude), down_longitude, longitude)) %>% 
#Option to omit rows without coordinates:
  filter(if_any(c("latitude", "longitude"), ~ !is.na(.)))

nrow(biodata.coord) 
#Note: we are missing lat/long values still for most sites 

head(biodata.coord)

View(biodata.coord)

#Check agencies that have coordinate data
#Note missing coordinate data for Browns Creek
levels(factor(biodata.coord$agency))


```

Transform coordinates in WGS84 to NAD83:

```{r}

#Create a spatial object from coodinates:

biodata_sf<-st_as_sf(biodata.coord, coords=c("longitude","latitude"), crs=4326) #4326 is code for WGS84

#Transform to NAD83

biodata_NAD83<-st_transform(biodata_sf, crs=4269)

```

# Map of biomonitoring locations

Plot data to see if it looks OK:

(and add some other layers for context)

basemap examples: <https://r-tmap.github.io/tmap/articles/tmap_vv.html>

map colors from color brewer: <https://colorbrewer2.org/>

```{r}

#read in county boundaries for MN using tigris package
counties <-counties("Minnesota")
head(counties)
names(counties)

st_crs(counties) #check coordinate reference system 


#Create list of counties in TC Metro 
Metro.List<-c("Anoka", "Carver", "Dakota", "Hennepin", "Ramsey", "Scott", "Washington")

#Filter county shapefile to TC Metro only 
TC.Metro.counties<-counties %>% 
  filter(NAME %in% Metro.List)


#Plot data to see if it looks OK:
names(biodata_NAD83)

#Add in NHD streams layer:
#First dissolve the boundaries between TC counties to create a single contiguous polygon for the area of interest (AOI)
TC_contig <- st_union(TC.Metro.counties)

NHD_TC<-get_nhdplus(
  AOI = TC_contig)

#order of agencies to show in plot
#put MPCA first so it doesn't cover up all the other agencies (where there is overlap) OR
#put MPCA last so you can see what data you have in addition to MPCA data

levels(factor(biodata_NAD83$agency))
class(biodata_NAD83$agency)

biodata_NAD83$agency<-as.factor(biodata_NAD83$agency)
levels(biodata_NAD83$agency)

#Make nicer labels:
#biodata_NAD83$agency<-factor(biodata_NAD83$agency, levels=c("Anoka Conservation District", "Carver County", "MCES", "Riley Purgatory", "Scott County WMO", "Washington Conservation District", "SHEP", "VSMP", "MPCA"))

#Plot the map
#different colors for each agency, need to specify number of colors below 

tmap_mode("view")


#map colors from color brewer
#Note: you must have same # of colors as you do agencies
color.palette<-c('#542788','#1f78b4','#b2df8a','#33a02c','#fb9a99','#e31a1c','#fdbf6f','#ff7f00','#cab2d6', '#fee090')
length(color.palette)  


Biodata.map<-
tm_shape(TC.Metro.counties)+
  tm_borders()+
tm_shape(NHD_TC)+
  tm_lines(col="#2c7bb6", lwd=0.5)+
tm_shape(biodata_NAD83)+
#sites sampled by agencies other than MPCA:
  #tm_shape(biodata_NAD83 %>% filter(!agency=="MPCA"))+
  tm_dots(fill="agency", fill.scale=tm_scale(values=color.palette), tm_legend(title="Agency"))+
  tm_layout(
    legend.position = c("right", "top"),  # or use c(0.95, 0.95) for precise control
    legend.bg.color = "white",
    legend.bg.alpha = 0.7,
    inner.margins = c(0.05,0.05,0.05,1.0)  # adds space around the map to avoid overlaps
  )
Biodata.map

#save map
tmap_save(Biodata.map, filename = paste(maps.dir, "/TC_biodata_map_9_16_25.png", sep=""), dpi = 300, width = 5.5, height = 6)

```

Number of unique sampling sites:

```{r}
names(biodata.coord)

biodata.coord %>% 
  dplyr::select(latitude, longitude) %>% 
  distinct() %>% 
  count()
```

### Write biodata (with corrected coordinates) to file

Options to write it to a shapefile that you can load directly into ArcGIS, or as a csv file.

```{r}

#Write as CSV file to UrbanBiomonitoring R Project: 
write.csv(biodata_NAD83 %>% st_drop_geometry(), paste0(dir.output, "/TC_final_combined_data.csv"), row.names=FALSE)

#Write as a shapefile
#NOTE: there may be some formatting issues with site names - may need to fix 
st_write(biodata_NAD83, paste0(gis.dir, "/TC_final_combined_data.shp"), append=FALSE)
```

# Overlay biodata with water chem data

## Overlay with Claire's chloride sites

Do these chloride sites overlap with biomonitoring data?

Read in Claire's chloride sampling locations:

```{r}

CL.sites<-read_excel(paste(dir.chem, "/Copy_of_Claire_Bass_Chloride_Sampling_Sites.xlsx", sep=""))
head(CL.sites)

#Convert to shapefile for mapping:
#Drop sites with missing coordinates
#Set coordinate ref system to NAD83

Cl_sf<-st_as_sf(CL.sites %>% na.omit(), coords=c("Longitude", "Latitude"), crs=4269)

#Quick map to check
plot(Cl_sf %>% st_geometry())
```

Read in biomonitoring sites (if not already loaded)

```{r}
biodata_NAD83<-st_read(paste(gis.dir, "/TC_final_combined_data.shp", sep=""))

plot(biodata_NAD83 %>% st_geometry())
```

Plot chloride data on top of biomonitoring sites

Question: what should count as a shared site? Ideas:

-   within 100m (or some distance)

-   on same streamreach of NHD

-   Cl sample upstream of biosample

```{r}


tmap_mode("view")

#map colors from color brewer
color.palette<-c('#a6cee3','#1f78b4','#b2df8a','#33a02c','#fb9a99','#e31a1c','#fdbf6f','#ff7f00','#cab2d6')
  

tm_shape(TC.Metro.counties)+
  tm_borders()+
tm_shape(NHD_TC)+
  tm_lines(col="darkturquoise", lwd=0.5)+
tm_shape(biodata_NAD83)+
  tm_dots(fill="agency", fill.scale=tm_scale(values=color.palette), tm_legend(title="Agency"), size=0.5)+
tm_shape(Cl_sf)+ #add Claire chloride locations
  tm_symbols(col="#f1a340", shape=1, size=0.5, lwd=0.5)





```

## Overlay with water chem data from public repositories

water chem data from script: 'module_TC_WQ_data_compilation_script.Rmd'\
and also show streamlines - from NHD

```{r}



tmap_mode("plot")

TC.map<-tm_shape(TC.Metro.counties)+
  tm_borders()+
tm_shape(NHD_TC)+
  tm_lines(col="#2c7bb6", lwd=0.5)+
tm_shape(biodata_NAD83)+
  tm_dots(fill="#998ec3", size=0.5)+
tm_shape(WQP_sites_TC)+ #add water chem locations from WQP
  tm_symbols(col="#f1a340", type="circle", size=0.5)+
  tm_add_legend(type="symbol", col=c("#998ec3", "#f1a340"),
                fill=c("#998ec3", "white"), border.col=c("#998ec3", "#f1a340"), labels=c("Biomonitoring samples", "Water chemistry samples"), shape=c(19,21))+
   tm_layout(
    legend.position = c("right", "top"),  # or use c(0.95, 0.95) for precise control
    #legend.bg.color = "white",
    legend.bg.alpha = 0.7,
    inner.margins = c(0.05,0.05,0.05,0.75)  # adds space around the map to avoid overlaps
  )
TC.map

#save map
tmap_save(TC.map, filename = paste(maps.dir, "/TC_biodata_waterchem_map_5_20_25.png", sep=""), dpi = 300, width = 5, height = 6)
```

# Look at data collection over time

for UCOWR presentation

sites sampled repeatedly over time

```{r}

#check date range by site

names(biodata.coord)

View(biodata.coord %>% 
  dplyr::select(original_id, date, agency) %>% 
  distinct() %>% 
  group_by(original_id, agency) %>% 
  count() %>%
  ungroup() %>% 
  filter(n>=2))

biodata.coord %>% 
  filter(original_id=="Rum River") %>% 
  dplyr::select(date) %>% 
  distinct()
  

```
