---
title: "module_MPCA_data_processing"
author: "Christy Dolph"
date: "2026-04-09"
output: html_document
editor_options: 
  chunk_output_type: console
---

# Overview

This script provides initial processing of biomonitoring datasets received from the Minnesota Pollution Control Agency. Excel input files used below were exported from the MPCA Access Database.

# Workspace Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Clean up workspace and load packages (or install if necessary)
#rm(list=ls())

want <- c("readxl", "writexl", "tidyverse", "zoo", 
          "sf", "tmap", "purrr", "fs", "janitor")
          
          need <- want[!(want %in% installed.packages()[,"Package"])]
if (length(need)) install.packages(need)
lapply(want, function(i) require(i, character.only=TRUE))
rm(want, need)

#Overall project directory:
root.dir <- getwd() 
root.dir


#Assign Directories 


#Output Data Directory for processed data:
output.dir <- paste0(root.dir, "/Data/cleaned_input_data") 
output.dir

#Raw input data directory:
mpca.dir<-paste0(root.dir, "/Data/biodata_raw/MPCA")
```

# Read in biodata

## i. Macroinvertebrates

MPCA macroinvertebrate data includes data collected by different methods, described in 'Protocol' column. I will select only records collected via 'QMH' = "Qualitative Multi-Habitat", which is 85% of all records.

-Keep life stage as larval and adult records may need to be combined

```{r}

#look at all files:
list.files(mpca.dir)

#read in macroinvertbrate data
mpca.bugs<-read_excel(paste0(mpca.dir, "/DataInverts_WriteOff.xlsx"))

names(mpca.bugs)

#Check for duplicates
View(mpca.bugs %>% janitor::get_dupes()) #no duplicates

#Read in taxonomic info
tax<-read_excel(paste0(mpca.dir, "/TaxaListInverts_WriteOff.xlsx"))

names(tax)

mpca.tax<-merge(mpca.bugs, tax %>% dplyr::select(-c(Name1, Name2, CommonName)), by=c("tsn"))

#Check number of rows lost:
nrow(mpca.bugs)-nrow(mpca.tax)

#Check for duplicates
View(mpca.bugs %>% janitor::get_dupes()) #no duplicates

#very small number of rows (3) lost after merging with taxa table

names(mpca.tax)

```

## ii. Fish

```{r}

#look at all files:
list.files(mpca.dir)

#read in macroinvertbrate data
mpca.fish<-read_excel(paste0(mpca.dir, "/DataFish_WriteOff.xlsx"))

names(mpca.fish)

#Check for duplicates
mpca.fish %>% janitor::get_dupes() #no duplicates

#Read in taxonomic info
tax.fish<-read_excel(paste0(mpca.dir, "/TaxaListFish_WriteOff.xlsx"))

names(tax.fish)

mpca.tax.fish<-merge(mpca.fish, tax.fish %>% dplyr::select(-c(Name1, Name2, CommonName)), by=c("tsn"))

#Check number of rows lost:
nrow(mpca.fish)-nrow(mpca.tax.fish) #should be 0

```

# Read in & merge site info to taxonomic data

```{r}

list.files(mpca.dir)

#Read in site and visit info and merge
#(remove 1 duplicate from site table)

sites<-read_excel(paste0(mpca.dir, "/Station_WriteOff.xlsx"))

visits<-read_excel(paste0(mpca.dir, "/Visit_Writeoff.xlsx"))
nrow(visits)

site.visit<-merge(sites %>% distinct(), visits, by=c("FieldNum"))
nrow(site.visit)

names(site.visit)

visit.fish<-read_excel(paste0(mpca.dir, "/VisitFish_WriteOff.xlsx"))
names(visit.fish)


#Add info about gear type from VisitFish table:
site.visit.fish<-left_join(site.visit, visit.fish, by=c("VisitNum"))
nrow(site.visit.fish)

#View(site.visit.fish)

```

## i. Merge site info to macroinvertebrate data

```{r}


#merge station info to bug data
#first fix any potential formatting issues with VisitNum and FieldNum

mpca.info<-merge(
  
  mpca.tax %>% 
    mutate(across(c(FieldNum, VisitNum), ~ toupper(as.character(.x)))) %>% 
    dplyr::select(-c(DateEntered, Comments.x, Comments.y)),
  
  site.visit %>% 
    mutate(across(c(FieldNum, VisitNum), ~ toupper(as.character(.x)))) %>% 
    filter(VisitType=="bugs") %>%  #select macroinvert visits only 
    dplyr::select(-c(DateEntered, Comments.x)),
  
  , by=c("VisitNum", "FieldNum"))
names(mpca.info)

#Check number of rows lost/gained
nrow(mpca.tax)-nrow(mpca.info) 

#Check for duplicates
mpca.info %>% janitor::get_dupes()
#no duplicates

#ID differences - find rows that didn't join

missing<-anti_join(
  mpca.tax %>% 
    mutate(across(c(FieldNum, VisitNum), ~ toupper(as.character(.x)))) %>% 
    dplyr::select(-c(DateEntered, Comments.x, Comments.y)),
  
  site.visit %>% 
    mutate(across(c(FieldNum, VisitNum), ~ toupper(as.character(.x)))) %>% 
    dplyr::select(-c(DateEntered, Comments.x)),
  
  , by=c("VisitNum", "FieldNum"))

View(missing)

levels(factor(missing$VisitNum))

#OK, there are some discrepancies in the original data for these rows - the FieldNum in the DataInverts file doesn't match the FieldNum in the Visit and VisitInvert file for these Visit Numbers - 9 visit numbers in total 

#I think OK to drop these data as record keeping errors 


```

## ii. Merge site info to fish data

```{r}

#merge station info to bug data
#first fix any potential formatting issues with VisitNum and FieldNum

names(mpca.tax.fish)

mpca.info.fish<-merge(
  
  mpca.tax.fish %>% 
    mutate(across(c(FieldNum, VisitNum), ~ toupper(as.character(.x)))) %>% 
    dplyr::select(-c(DateEntered, CNCode.x, CNCode.y)),
  
  site.visit.fish %>% 
    mutate(across(c(FieldNum, VisitNum), ~ toupper(as.character(.x)))) %>% 
    filter(VisitType=="fish") %>%  #select macroinvert visits only 
    dplyr::select(-c(DateEntered, Comments.x)),
  
  , by=c("VisitNum", "FieldNum"))
names(mpca.info.fish)

#Check number of rows lost/gained
nrow(mpca.tax.fish)-nrow(mpca.info.fish) 

#Check for duplicates
mpca.info.fish %>% janitor::get_dupes()
#no duplicates


```

# Number of replicates

Create replicate column for more than one visit for site/date combo

Add a seqential number for each unique Visit within a Site/Date group

```{r}

#Macroinvertebrates:
names(mpca.info)

replicates<-mpca.info %>% 
  dplyr::select(FieldNum, VisitDate, VisitNum) %>% 
  distinct() %>% 
  group_by(FieldNum, VisitDate) %>%  
  mutate(replicate = row_number())

#replicates

View(replicates %>% filter(replicate>1))
View(replicates %>% filter(FieldNum=="96SC034"))

mpca.info.rep<-merge(mpca.info, replicates, by=c("FieldNum", "VisitDate", "VisitNum"))

names(mpca.info.rep)

nrow(mpca.info)
nrow(mpca.info.rep)

#Fish:

names(mpca.info.fish)

replicates<-mpca.info.fish %>% 
  dplyr::select(FieldNum, VisitDate, VisitNum) %>% 
  distinct() %>% 
  group_by(FieldNum, VisitDate) %>%  
  mutate(replicate = row_number())

#replicates

#View(replicates %>% filter(replicate>1))

mpca.info.rep.fish<-merge(mpca.info.fish, replicates, by=c("FieldNum", "VisitDate", "VisitNum"))

names(mpca.info.rep.fish)

nrow(mpca.info.fish)
nrow(mpca.info.rep.fish)

```

# Subset columns and write to file

## i. Macroinvertebrates

```{r}
#Target columns needed to combine with other datasets 

names(mpca.info.rep)

target_cols<-c("Original_ID",	"Collection_Date", "Season",	"Year",	"Replicate",	"Agency",	"Collected_by",	"Class", "Subclass", "Order", "Family",	"Genus", "Species", "Original_Taxon",	"Number",	"Notes", "TSN",	"Site", "Latitude", "Longitude", "Northing", "Easting", "total_replicates")

#Subset to needed columns and QMH samples
mpca.all<- mpca.info.rep %>% 
  filter(Protocol=="QMH") %>%
  dplyr::rename(Original_ID=FieldNum, Number=Count, Original_Taxon=Taxon, Latitude=LAT8xDD, Longitude=LON8xDD, Collection_Date=VisitDate, Year=VisitYear) %>% 
  mutate(Agency="MPCA") %>% 
  mutate(Collection_Date=as.Date(Collection_Date, format="%m/%d/%Y")) %>% 
  dplyr::select(Original_ID, Collection_Date, Year, Agency, Family, Genus, Original_Taxon, Number, tsn, Latitude, Longitude, LifeStage, replicate)

names(mpca.all)

#View(mpca.all)

#Write to file

write.csv(mpca.all, paste0(output.dir, "/mpca_inverts_compiled.csv"), row.names=FALSE)
```

## ii. Fish

```{r}


#Target columns needed to combine with other datasets 

names(mpca.info.rep.fish)

target_cols<-c("Original_ID",	"Collection_Date", "Season",	"Year",	"Replicate",	"Agency",	"Collected_by",	"Class", "Subclass", "Order", "Family",	"Genus", "Species", "Original_Taxon",	"Number",	"Notes", "TSN",	"Site", "Latitude", "Longitude", "Northing", "Easting", "total_replicates")

#Subset to needed columns and QMH samples
mpca.all.fish<- mpca.info.rep.fish %>% 
  dplyr::rename(Original_ID=FieldNum, Latitude=LAT8xDD, Longitude=LON8xDD, Collection_Date=VisitDate, Year=VisitYear, Genus=Name1, Species=Name2) %>% 
  mutate(FullScientificName = paste(Genus, Species, sep = " ")) |> 
  mutate(Agency="MPCA") %>% 
  mutate(Collection_Date=as.Date(Collection_Date, format="%m/%d/%Y")) %>% 
  dplyr::select(Original_ID, Collection_Date, Year, Agency,Genus,Species, FullScientificName, CommonName, tsn, Number, Latitude, Longitude, replicate, GearType)

names(mpca.all.fish)

#View(mpca.all.fish)

#Write to file

write.csv(mpca.all.fish, paste0(output.dir, "/mpca_fish_compiled.csv"), row.names=FALSE)

```

# Data summary

```{r}

#Date range

mpca.all %>% dplyr::summarise(mindate=min(Collection_Date), maxdate=max(Collection_Date))

#Number of unique sites
mpca.all %>% dplyr::select(Latitude, Longitude) %>% distinct() %>% nrow()

#View taxonomy

View(mpca.all %>% dplyr::select(Family, Genus, Original_Taxon))
```
