---
title: "module_fish_OTUs"
author: "Christy Dolph"
date: "2026-10-08"
output: html_document
editor_options: 
  chunk_output_type: console
---

# Overview

This script checks and resolves operational taxonomic units (OTUs) for fish taxa and creates a rarified relative abundance dataset at several different taxonomic resolutions (family, genus, and mixed ID).

Options:

-throw out hybrids

-throw out coarser than species IDs

-for USGS/EPA, if contains ".x." list as hybrid with genus name

\-

# Set Up Workspace

```{r}
knitr::opts_chunk$set(echo = TRUE)

# Clean up workspace and load packages (or install if necessary)
rm(list=ls())

want <- c("readxl", "writexl", "tidyverse", "zoo", 
          "tigris", "sf", "tmap", "purrr", "fs", "pak", "fuzzyjoin", "data.table", "stringr", "taxize")
          
          need <- want[!(want %in% installed.packages()[,"Package"])]
if (length(need)) install.packages(need)
lapply(want, function(i) require(i, character.only=TRUE))
rm(want, need)


#Designate directoriers:


#Overall project directory:
root.dir <- getwd() 
root.dir

#Input directory for raw/unprocessed data
raw.dir<-paste0(root.dir, "/Data/biodata_raw")
raw.dir

#Output Data Directory for processed data:
input.dir <- paste0(root.dir, "/Data/cleaned_input_data") 
input.dir

#Output directory for finalized taxonomic datasets
taxa.dir<-paste0(root.dir, "/Data/processed_taxonomic_datasets")

#Output directory for tables/figures
fig.dir<-paste0(root.dir, "/Data/Figures_Tables")
```

# Load in all sources of data

- MPCA

- finsync (USGS and EPA data, preprocessed by Mahon et al 2024)

Note that all data are raw counts.

```{r}

list.files(input.dir)

mpca<-as.data.frame(fread(paste0(input.dir, "./mpca_fish_compiled.csv")))
names(mpca)
#View(mpca)

nat<-as.data.frame(fread(paste0(input.dir, "./USGS_EPA_Fish_MN_long_finsync.csv")))
names(nat)
#View(nat)


```

QC check - check for duplicates in individual datasets

```{r}

##MPCA:
mpca %>% janitor::get_dupes()
#no duplicates


#USGS/EPA:
nat %>% 
    filter(Number>0) %>% 
    janitor::get_dupes()
  
#No dupes across all columns, but: 

#When I subset columns to those I use in aggregated data, get duplicates:

nat.dup.check<-
nat %>% 
 mutate(Date=as.Date(CollectionDate, format="%Y-%m-%d")) %>% #format date
  dplyr::rename(Original_ID=SiteNumber, Year=CollectionYear, FullScientificName=Name, Latitude=Latitude_dd, Longitude=Longitude_dd, GearType=SampleMethod) %>% #format column names
    mutate(replicate=1) %>%  #add replicate column (no site/date replicates in USGS/EPA dataset)
    dplyr::select(Original_ID, Date, Year, Agency, FullScientificName, Number, Latitude, Longitude, replicate, GearType, COMID) %>% 
  filter(Number>0) %>% 
    janitor::get_dupes()



```

# Clean up data to match across datasets

For MPCA, how many records are genus only (so species ID)

```{r}

(mpca |> filter(is.na(Species)) |> nrow())/nrow(mpca) #.5% of records are genus only ID - can exclude

mpca2<-mpca |> filter(!is.na(Species)) |> 
  mutate(Date=as.Date(Collection_Date, format="%Y-%m-%d")) |> 
  dplyr::select(-c(Collection_Date))

```

For USGS/USEPA, format columns to match MPCA dataset

```{r}


#Cleaner USGS/USEPA data:
nat2<-nat%>% 
  mutate(Date=as.Date(CollectionDate, format="%Y-%m-%d")) %>% #format date
  dplyr::rename(Original_ID=SiteNumber, Year=CollectionYear, FullScientificName=Name, Latitude=Latitude_dd, Longitude=Longitude_dd, GearType=SampleMethod) %>% #format column names
  mutate(replicate=1) %>%  #add replicate column (no site/date replicates in USGS/EPA dataset)
  dplyr::select(Original_ID, Date, Year, Agency, FullScientificName, Number, Latitude, Longitude, replicate, GearType, COMID) |> 
  mutate(
    FullScientificName = if_else(
      str_detect(FullScientificName, fixed(".x.")),
      str_replace(FullScientificName, "\\..*", " hybrid"),
      FullScientificName)) |>  #replace any hybrid with "hybrid"
  mutate(FullScientificName = str_replace_all(FullScientificName, "\\.", " ")) |> #format scientific name to match MPCA data
  mutate(Genus = str_extract(FullScientificName, "^[^ ]+")) |>  #Extract genus name
  mutate(Species = str_extract(FullScientificName, "(?<= )[^ ]+")) |> 
  filter(Number>0) #remove 0 counts

head(nat2)

levels(factor(mpca2$FullScientificName))
levels(factor(nat2$FullScientificName))

#Check names that don't match across datasets
#find names in col A that are not in col B
setdiff(mpca2$FullScientificName, nat2$FullScientificName)

setdiff(nat2$FullScientificName,mpca2$FullScientificName) #mpca dataset much more extensive - many more species 

```

Merge data sources to same columns.

```{r}

names(mpca2)
names(nat2)


all.fish<-bind_rows(
  mpca2,
  nat2)
  

nrow(all.fish)

View(all.fish)

#Fill out missing data if names match

#lookup table to match common names from mpca dataset to scientific names
lookup <- all.fish %>%
  filter(!is.na(FullScientificName), !is.na(CommonName)) %>%
  distinct(FullScientificName, Name_filled = CommonName)
lookup

# 2. Join and fill missing values
all.fish2 <- all.fish %>%
  left_join(lookup, by = "FullScientificName") %>%
  mutate(CommonName = coalesce(CommonName, Name_filled)) %>%
  select(-Name_filled)

View(all.fish2 |> filter(!Agency=="MPCA"))

#check for any remaining missing names - species in USGS/USEPA datasets that are not in MPCA dataset 

missing.common<-all.fish2 |> filter(is.na(CommonName))  |> dplyr::select(FullScientificName) |> distinct()

nrow(all.fish2)

CommonName.missing<-c(
"Eastern blacknose dace",
"Rosyface shiner",
"Allegheny pearl dace",
"American pickerel",
"American brook lamprey",
"Slender chub hybrid",
"Silver shiner",
"Blue catfish",
"Greenside darter",
"Orangethroat darter",
"Dusky darter")

missing.lookup<-cbind(missing.common, CommonName.missing)

missing.lookup

#Join this table for remaining missing common names of species in USGS/USEPA but not in mpca dataset:

all.fish3<-all.fish2 %>%
  left_join(missing.lookup, by = "FullScientificName") %>%
  mutate(CommonName = coalesce(CommonName, CommonName.missing)) %>%
  select(-CommonName.missing) |> 
  mutate(CommonName=str_to_sentence(CommonName)) |> #capitalize all common names
#remove NA values for number - checked and these were NA in original mpca data - no count
  filter(!is.na(Number))

View(all.fish3 |> filter(FullScientificName %in% missing.common$FullScientificName))

View(all.fish3)

#any missing data?
View(all.fish3 |>  dplyr::select(-c(tsn,COMID))|> filter(if_any(everything(), is.na)))

#there is a small number of rows of MPCA data for which GearType is not available, I think likely OK to assume electroshocking

```

Quick map check:

```{r}

all.fish_sf<-st_as_sf(all.fish3, coords=c("Longitude", "Latitude"), crs=4269, remove=FALSE)

names(all.fish_sf)

plot(all.fish_sf$geometry)

```

# Check for duplicates across datasets

```{r}

all.fish3 |> dplyr::select(Latitude, Longitude, Date, Agency) |> distinct() |> group_by(Latitude, Longitude, Date)|> count() |> filter(n>1)

```

# Calculate Relative Abundance

```{r}

#Sum count by site/date

all.fish4<-all.fish3 |>
  group_by(Original_ID, Date) |> 
  mutate(TotalCount=sum(Number)) |> 
  ungroup() |> 
  mutate(Relative_Abund=Number/TotalCount)

View(all.fish4)

#save to file
#exclude tsn and COMID for now (may want back later)
write_xlsx(all.fish4 |> dplyr::select(-c(tsn, COMID)), paste0(taxa.dir, "/Fish_relative_abundance.xlsx"))
```
