---
title: "WQ_data_compilation"
author: "Christy Dolph"
date: "2025-04-24"
output: html_document
editor_options: 
  chunk_output_type: inline
---

# Overview

The goal of this script is to assemble water chemistry data for streams in the Twin Cities Metro. Data is assembled from multiple public data respositories including:

-   Water Quality Portal

-   MetCouncil EIMS portal

Ultimately we are planning to match this water chemistry data to locations where biological data (macroinvertebrate diversity & abundancy) has also been collected.

## Workspace setup

Load required packages and create directories

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

#Clean up workspace and load packages (or install if necessary)

#rm(list=ls())
want <- c('here', 'data.table', 'tidyverse','tidyr', 'readxl','readr', 'purrr', 'writexl', 'mapview', 'sf', 'terra', 'tmap','geodata', 'RSelenium', 'rvest', 'httr', 'dataRetrieval', 'DT', 'FedData', 'raster', 'rgdal', 'nhdplusTools', 'tigris', 'janitor', 'oce', 'utm2lonlat')
          
          need <- want[!(want %in% installed.packages()[,"Package"])]
if (length(need)) install.packages(need)
lapply(want, function(i) require(i, character.only=TRUE))
rm(want, need)


#Set up working directories

#Create input and output folder destinations, if they don't already exist:
dir.create(file.path(getwd(), "Water_chemistry_datasets"))


dir <- list()
root.dir <- getwd() #Set Main Directory one level up
list.files(root.dir)

#Data Directory:
data.dir <- paste0(root.dir, "/Data/biodata_raw/TC_Metro/Compiled_data_clean") 
data.dir

#GIS directory
gis.dir<-paste0(root.dir, "/Data/GIS")
gis.dir

#Map images directory
maps.dir<-paste0(root.dir, "/Maps")
maps.dir

#Water chem directory
dir.chem <- paste0(root.dir, "/Data/water_chemistry")
dir.chem

#Output data
dir.output<-paste0(root.dir, "/Data/biodata_processed")
dir.output
```

# Water Quality Portal

Water Quality Portal User Guide: <https://www.waterqualitydata.us/portal_userguide/>

Look at available list of water quality parameters. For now prioritize the following:

-   Chloride

-   Nitrogen

-   Phosphorus

-   Suspended sediment

-   Lead

-   PCBs

```{r}
#Read in all parameter codes available in NWIS
pcode <- readNWISpCode("all")

#Look at all codes:
#View(pcode)

#Check categories of codes
levels(factor(pcode$parameter_group_nm))

#Pull out water chemistry related codes of interest

#want suspended sediment but not necessarily other rows with "sediment"
  
ChemCds<-pcode %>% 
  filter(grepl("nitrogen|nitrate|phosphorus|chloride|lead|PCBs|`Suspended sediment`", parameter_nm, ignore.case = TRUE)) %>% 
  filter(!parameter_group_nm=="Radiochemical") #filter out radiochem

View(ChemCds)

names(ChemCds)

#Option to view available P parameters in a table (as HTML table using DT package):
datatable(ChemCds[, c("parameter_cd", "parameter_nm", "parameter_units")],
  rownames = FALSE, options = list(pageLength = 4))
```

## Further narrow down to parameter codes of interest:

These parameters were selecting by viewing the 'ChemCds' table (see above chunk), and handpicking parameters.

Some of these parameters seem to be redundant? Need to figure out how they are different and if some of them can be combined. Possible reasons:

-   different units

-   different lab/analytical methods

-   some are estimates from regression, vs sampled directly

Notes:

-   Pulling parameter codes for water, and in some cases bed sediment.

-   I am not pulling chemistry for different suspended/bed sediment size fractions, but this is available.

-   There are a lot of codes for specific metals chemistry in bed sediment (full metals workup) that I have not included here (only including lead).

-   I excluded parameter codes for loads (ie, lbs per day).

-   Most water chemistry has filtered and unfiltered option, need to figure out which you want.

```{r}

parameter_code=c('00600',
'00602',
'00605',
'00607',
'00608',
'00610',
'00613',
'00615',
'00618',
'00620',
'00630',
'00631',
'00665',
'00666',
'00667',
'00668',
'00670',
'00671',
'00673',
'00674',
'00676',
'00940',
'01049',
'01050',
'01051',
'01052',
'01072',
'01382',
'01388',
'01407',
'01473',
'04116',
'29836',
'30272',
'30291',
'30292',
'30335',
'30362',
'39516',
'39517',
'39518',
'39519',
'49567',
'49568',
'51185',
'51288',
'52286',
'52298',
'52635',
'52683',
'62854',
'62855',
'64832',
'65043',
'70387',
'70507',
'71850',
'71851',
'71886',
'71887',
'71888',
'90869',
'91001',
'91003',
'91004',
'99117',
'99120',
'99121',
'99123',
'99124',
'99125',
'99130',
'99133',
'99220',
'99404',
'99410',
'99411',
'99412',
'99414',
'99415',
'99416',
'99435',
'99889',
'99891',
'99892',
'99893',
'99894')

```

## Pull out WQP data for streams in MN

<https://cran.r-project.org/web/packages/dataRetrieval/vignettes/dataRetrieval.html>

"statecode" is FIPS state codes: <https://www.waterqualitydata.us/portal_userguide/>

Minnesota FIPS code is 27

```{r}
 
#search by state

Water.chem<-readWQPdata(statecode=c("US:27"), 
          siteType="Stream",
          sampleMedia="Water",
          parameterCd=parameter_code)
View(Water.chem)
```

## ID site locations in study area where water chem was collected

```{r}
#get site info
sites<-whatWQPsites(statecode=c("US:27"), 
          siteType="Stream",
          sampleMedia="Water",
          parameterCd=parameter_code)
#View(sites)
```

## Create spatial object and plot map

```{r}
names(sites)
WQP_sites_sf<-st_as_sf(sites, coords=c("LongitudeMeasure", "LatitudeMeasure"), crs=4326)


#read in USA state boundaries using tigris package
states<-states()
states_sf <- st_as_sf(states) #convert to sf object
st_bbox(states_sf) #bounding box
states_sf

st_crs(states_sf) #check reference system

#Project riparian TP samples in NAD83 to match state boundaries
WQP_sites_NAD83<-st_transform(WQP_sites_sf, crs=st_crs(states_sf))

#Plot riparian TP soil sample locations
tmap_mode("view")

tm_shape(WQP_sites_NAD83) +
  tm_dots()+
  tm_shape(states_sf)+
  tm_borders()
```

### Limit sites to Twin Cities Metro Area

```{r}
#read in county boundaries for MN using tigris package
counties <-counties("Minnesota")
head(counties)
names(counties)

st_crs(counties) #check coordinate reference system 


#Create list of counties in TC Metro 
Metro.List<-c("Anoka", "Carver", "Dakota", "Hennepin", "Ramsey", "Scott", "Washington")

#Filter county shapefile to TC Metro only 
TC.Metro.counties<-counties %>% 
  filter(NAME %in% Metro.List)

#Clip water chem data to metro
WQP_sites_TC<-st_intersection(WQP_sites_NAD83, TC.Metro.counties)

#Write to file so can intersect with biomonitoring data
st_write(WQP_sites_TC, paste(gis.dir, "/WQP_waterchem_TCMetro.shp", sep=""), append=FALSE)

#Plot data to see if it looks OK:

tm_shape(TC.Metro.counties)+
  tm_borders()+
tm_shape(WQP_sites_TC)+
  tm_dots(fill="orange")
```

Clip WQP points to TC Metro:

```{r}
WQP.TCMetro.sites<-st_intersection(WQP_sites_NAD83, TC.Metro.counties)

#Number of TC Metro area sites:
nrow(WQP.TCMetro.sites)

tm_shape(TC.Metro.counties)+
  tm_borders()+
tm_shape(WQP.TCMetro.sites)+
  tm_dots()

```

## Merge water chem data to site ID

Site ID is "MonitoringLocationIdentifier"

```{r}
names(sites)

names(Water.chem)

#For all of MN, merge water chem to site info, leaving out redundant info between tables: 
WQP.chem.sites<-merge(sites, (Water.chem %>% dplyr::select(!c(OrganizationIdentifier, OrganizationFormalName, ProviderName))), by=c("MonitoringLocationIdentifier"))
names(WQP.chem.sites)

#Merge to detailed information for water chem parameters
ChemCds<-ChemCds %>% 
  dplyr::mutate(CharacteristicName=srsname) %>% 
  dplyr::mutate(USGSPCode=parameter_cd)
View(ChemCds)

nrow(WQP.chem.sites)

WQP.chem.sites.specific<-merge(WQP.chem.sites, ChemCds %>% dplyr::select(!c(CharacteristicName)), by=c("USGSPCode"))
nrow(WQP.chem.sites.specific)

names(WQP.chem.sites.specific)
#View(WQP.chem.sites.specific)

##############################################
#For TC Metro:
#ID Metro Sites
names(WQP.TCMetro.sites)

Metro.sites<-WQP.TCMetro.sites %>% 
  st_drop_geometry %>% 
  dplyr::select(MonitoringLocationIdentifier) %>%
  distinct() %>% 
  as.list()

#Water chem for Metro sites:
WQP.chem.TCMetro<-WQP.chem.sites.specific %>% 
  filter(MonitoringLocationIdentifier %in% Metro.sites$MonitoringLocationIdentifier)

nrow(WQP.chem.TCMetro)

```

## Examine WQP water chemistry data

Look at how much data there is by parameter and site.

This dataset contains repeated measurements over time and space.

Notes:

-   Be careful with "CharacteristicName" as this is a very general category that lumps together different types of parameters

```{r}
names(WQP.chem.TCMetro)


#Look at number of samples collected by each site and parameter:
View(WQP.chem.TCMetro %>% 
  group_by(MonitoringLocationIdentifier, parameter_nm, ResultMeasure.MeasureUnitCode, USGSPCode) %>% 
  count())


#Check on dates of specific parameters


```

Number of sites sampled \>5 times for a given parameter

```{r}
names(WQP.chem.TCMetro)

View(WQP.chem.TCMetro %>% 
  group_by(MonitoringLocationIdentifier, parameter_nm, ResultMeasure.MeasureUnitCode, USGSPCode) %>% 
  count() %>% 
  filter(n>5) %>% 
  ungroup() %>% 
  dplyr::select(MonitoringLocationIdentifier) %>% 
  distinct())

#78 sites sampled >5 sites for at least 1 parameter
```

Most common parameters sampled across sites (but retain different parameter codes)

```{r}
View(WQP.chem.TCMetro %>% 
  group_by(parameter_nm, ResultMeasure.MeasureUnitCode, USGSPCode) %>% 
  count()) 
```

### Combining WQP parameters

Information on parameters: <https://help.waterdata.usgs.gov/codes-and-parameters/parameters>

One approach: combine constituents if they have the exact same parameter name, even if parameter codes differ (may be discrepancies in how these were measured??)

```{r}
#Look at specific constituent as an example

View(ChemCds %>% 
  filter(parameter_nm=="Phosphorus, water, unfiltered, milligrams per liter as phosphorus"))

#Look at most common parameters across sites, if parameter names match (disregarding codes)
View(WQP.chem.TCMetro %>% 
  group_by(parameter_nm, ResultMeasure.MeasureUnitCode) %>% 
  count()) 

```

# Getting Data from Met Council: EIMS

The Met Council data portal is called EIMS: <https://eims.metc.state.mn.us/AdvancedSearch>

Options for accessing data:

1.  Download data from the portal, and then load it into R - file name includes access date
2.  Use the web link for the search query to load it directly into R (tho note this creates additional formatting issues)

To pull data for all sites:

-   Search by watershed (huc8): select all 11 watersheds

-   site type = stream/river

-   leave date range open for all dates

-   parameters - select individually otherwise data is too large to download from the portal.

    -   Parameters selected: nutrients, inorganics, organics, physical

EIMS query for nutrients:

[https://eims.metc.state.mn.us/Download?startDate=01-01-0001&endDate=12-31-9999&parameterIds=2037;2038;2039;2163;2164;2221;2222;2223;2224;2225;2226;2172;2173;2231;2232;2233;2234;2235&siteTypes=River/Stream&majorWatersheds=Blue%20Earth%20River;Cannon%20River;Le%20Sueur%20River;Lower%20Minnesota%20River;Lower%20St.%20Croix%20River;Minnesota%20River%20-%20Mankato;Mississippi%20River%20-%20Lake%20Pepin;Mississippi%20River%20-%20Twin%20Cities;North%20Fork%20Crow%20River;Rum%20River;South%20Fork%20Crow%20River&shortSiteIds=\\\~0INn9wla-;E1FTFrJdB5DxCOBKJuD7C6MmJ7J_EbN0GFJ1EDEHJwF6EqN2JfFYFHDOJqFQNnElDEJlGBEvJeJnJ9JoFmCtEPFeFJFcEUJsEYDpDjBqDmB4DdJrDaBDDXMACjNKBmNZCLKODtNYJ8NWCiDDDfFwFNJ3MoCZCkCNCxDLDIDiM8M6NcNMDNJ6DuNLDcNTMZM5DMLxM4NJN9NONILTM9GSGQE0M2NSNAHLNHNPNCNRNDNN;&format=csv](https://eims.metc.state.mn.us/Download?startDate=01-01-0001&endDate=12-31-9999&parameterIds=2037;2038;2039;2163;2164;2221;2222;2223;2224;2225;2226;2172;2173;2231;2232;2233;2234;2235&siteTypes=River/Stream&majorWatersheds=Blue%20Earth%20River;Cannon%20River;Le%20Sueur%20River;Lower%20Minnesota%20River;Lower%20St.%20Croix%20River;Minnesota%20River%20-%20Mankato;Mississippi%20River%20-%20Lake%20Pepin;Mississippi%20River%20-%20Twin%20Cities;North%20Fork%20Crow%20River;Rum%20River;South%20Fork%20Crow%20River&shortSiteIds=~0INn9wla-;E1FTFrJdB5DxCOBKJuD7C6MmJ7J_EbN0GFJ1EDEHJwF6EqN2JfFYFHDOJqFQNnElDEJlGBEvJeJnJ9JoFmCtEPFeFJFcEUJsEYDpDjBqDmB4DdJrDaBDDXMACjNKBmNZCLKODtNYJ8NWCiDDDfFwFNJ3MoCZCkCNCxDLDIDiM8M6NcNMDNJ6DuNLDcNTMZM5DMLxM4NJN9NONILTM9GSGQE0M2NSNAHLNHNPNCNRNDNN;&format=csv){.uri}

EIMS query for inorganics:

[https://eims.metc.state.mn.us/Download?startDate=01-01-0001&endDate=12-31-9999&parameterIds=2074;2075;2151;2152;2192;2193;2205;2206;2056;2077;2085;2086;2087;2120;2121;2122;2202;2209;2210;2217;2218;2219;2036;2045;2054;2055;2072;2073;2099;2100;2276;2109;2110;2142;2144;2145;2148;2149;2153;2154;2155;2156;2158;2161;2162;2203;2204;2214;2215;2249;2250;2041;2042;2043;2044;2067;2111;2200;2201&siteTypes=River/Stream&majorWatersheds=Blue%20Earth%20River;Cannon%20River;Le%20Sueur%20River;Lower%20Minnesota%20River;Lower%20St.%20Croix%20River;Minnesota%20River%20-%20Mankato;Mississippi%20River%20-%20Lake%20Pepin;Mississippi%20River%20-%20Twin%20Cities;North%20Fork%20Crow%20River;Rum%20River;South%20Fork%20Crow%20River&shortSiteIds=\\\~nI0Nawl9-;BKCODOElEbDxC6B5JuEqEvD7E1EDGFEPN9N2N0NnFrJwDEDpDjDaDfFmMmFTFYCiNJMZF6JoBDB4DXDIJfJrCjCtNZCZBqCkNWDMDcJdEHFcCNCxMoDNJqJlGBDiDmKOJeJsJnE0J9J3FQJ1DtFNJ8J_DdLxNcJ7DuNMNNFJMANLEYFeEUGSFHFwGQM8NCNKM2M6NDM4NIM9NOLTNTM5J6NHNSDlNPNABmNYCLDDNRHL;&format=csv](https://eims.metc.state.mn.us/Download?startDate=01-01-0001&endDate=12-31-9999&parameterIds=2074;2075;2151;2152;2192;2193;2205;2206;2056;2077;2085;2086;2087;2120;2121;2122;2202;2209;2210;2217;2218;2219;2036;2045;2054;2055;2072;2073;2099;2100;2276;2109;2110;2142;2144;2145;2148;2149;2153;2154;2155;2156;2158;2161;2162;2203;2204;2214;2215;2249;2250;2041;2042;2043;2044;2067;2111;2200;2201&siteTypes=River/Stream&majorWatersheds=Blue%20Earth%20River;Cannon%20River;Le%20Sueur%20River;Lower%20Minnesota%20River;Lower%20St.%20Croix%20River;Minnesota%20River%20-%20Mankato;Mississippi%20River%20-%20Lake%20Pepin;Mississippi%20River%20-%20Twin%20Cities;North%20Fork%20Crow%20River;Rum%20River;South%20Fork%20Crow%20River&shortSiteIds=~nI0Nawl9-;BKCODOElEbDxC6B5JuEqEvD7E1EDGFEPN9N2N0NnFrJwDEDpDjDaDfFmMmFTFYCiNJMZF6JoBDB4DXDIJfJrCjCtNZCZBqCkNWDMDcJdEHFcCNCxMoDNJqJlGBDiDmKOJeJsJnE0J9J3FQJ1DtFNJ8J_DdLxNcJ7DuNMNNFJMANLEYFeEUGSFHFwGQM8NCNKM2M6NDM4NIM9NOLTNTM5J6NHNSDlNPNABmNYCLDDNRHL;&format=csv){.uri}

EIMS query for organics:

[https://eims.metc.state.mn.us/Download?startDate=01-01-0001&endDate=12-31-9999&parameterIds=2001;2002;2003;2004;2005;2006;2007;2010;2011;2016;2017;2018;2019;2020;2021;2023;2024;2028;2030;2033;2034;2040;2047;2048;2049;2050;2051;2052;2053;2057;2058;2059;2060;2068;2069;2071;2076;2089;2090;2091;2101;2113;2114;2116;2117;2118;2119;2129;2132;2133;2139;2140;2141;2143;2146;2150;2157;2159;2160;2165;2166;2167;2168;2169;2175;2176;2187;2188;2195;2213;2216;2227;2228;2230;2238;2241;2242;2247;2177;2178;2179;2180;2181;2182;2183;2229;2008;2009;2012;2013;2014;2015;2022;2025;2026;2027;2029;2031;2032;2035;2046;2070;2084;2088;2102;2112;2115;2124;2125;2126;2127;2128;2134;2136;2137;2138;2184;2237;2239&siteTypes=River/Stream&majorWatersheds=Blue%20Earth%20River;Cannon%20River;Le%20Sueur%20River;Lower%20Minnesota%20River;Lower%20St.%20Croix%20River;Minnesota%20River%20-%20Mankato;Mississippi%20River%20-%20Lake%20Pepin;Mississippi%20River%20-%20Twin%20Cities;North%20Fork%20Crow%20River;Rum%20River;South%20Fork%20Crow%20River&shortSiteIds=\\\~0nINwl-;EbD7DOBKC6DxB5E1EvCOEqN0MZElFYJwDtJnFrDdJlJ1EPGFFmF6JoFcEDFTDNKOJdDpGBFQFHJuMmDjJ9JeEHCiJsDfJfCkDIMoDaJqNWJ3NnNLN2N9DMNMJ7BqLxFwFeM6BDB4NZCxCZ;&format=csv](https://eims.metc.state.mn.us/Download?startDate=01-01-0001&endDate=12-31-9999&parameterIds=2001;2002;2003;2004;2005;2006;2007;2010;2011;2016;2017;2018;2019;2020;2021;2023;2024;2028;2030;2033;2034;2040;2047;2048;2049;2050;2051;2052;2053;2057;2058;2059;2060;2068;2069;2071;2076;2089;2090;2091;2101;2113;2114;2116;2117;2118;2119;2129;2132;2133;2139;2140;2141;2143;2146;2150;2157;2159;2160;2165;2166;2167;2168;2169;2175;2176;2187;2188;2195;2213;2216;2227;2228;2230;2238;2241;2242;2247;2177;2178;2179;2180;2181;2182;2183;2229;2008;2009;2012;2013;2014;2015;2022;2025;2026;2027;2029;2031;2032;2035;2046;2070;2084;2088;2102;2112;2115;2124;2125;2126;2127;2128;2134;2136;2137;2138;2184;2237;2239&siteTypes=River/Stream&majorWatersheds=Blue%20Earth%20River;Cannon%20River;Le%20Sueur%20River;Lower%20Minnesota%20River;Lower%20St.%20Croix%20River;Minnesota%20River%20-%20Mankato;Mississippi%20River%20-%20Lake%20Pepin;Mississippi%20River%20-%20Twin%20Cities;North%20Fork%20Crow%20River;Rum%20River;South%20Fork%20Crow%20River&shortSiteIds=~0nINwl-;EbD7DOBKC6DxB5E1EvCOEqN0MZElFYJwDtJnFrDdJlJ1EPGFFmF6JoFcEDFTDNKOJdDpGBFQFHJuMmDjJ9JeEHCiJsDfJfCkDIMoDaJqNWJ3NnNLN2N9DMNMJ7BqLxFwFeM6BDB4NZCxCZ;&format=csv){.uri}

EIMS query for inorganics:

[https://eims.metc.state.mn.us/Download?startDate=01-01-0001&endDate=12-31-9999&parameterIds=2107;2108;2135;2106;2190;2198;2240;2272;2243;2244;2245;2246;2191;2196;2061;2062;2063;2064;2065;2066;2078;2079;2080;2081;2082;2083;2103;2104;2105;2174;2207;2185;2186;2194;2211;2220;2248;2273;2197;2208;2212&siteTypes=River/Stream&majorWatersheds=Blue%20Earth%20River;Cannon%20River;Le%20Sueur%20River;Lower%20Minnesota%20River;Lower%20St.%20Croix%20River;Minnesota%20River%20-%20Mankato;Mississippi%20River%20-%20Lake%20Pepin;Mississippi%20River%20-%20Twin%20Cities;North%20Fork%20Crow%20River;Rum%20River;South%20Fork%20Crow%20River&shortSiteIds=\\\~0wnNIa-l9MLWV;EbEUElEPDxJuEqC6JdD7DOJ1COB5LTE1EvFrDIFNMAGFBKDECiDiB4FmGBFQFeFHNZDdF6FJEDCNFTKOJ3JnCtEYJ_JfJeDpJqLxFYJ9FwJ8DXJwDmFcNMCxDcJ6M8BqNKJlJsBDJ7DfCkNWNNCZJoMmEHDtMoM9DjM6DaMZE0DuM5NLNTNJM4NONnDNDMNINRN0N2BmDDNHCLNSNPNYJrCjNAHLN9NCNDNcM2GQGSN6NqDL;&format=csv](https://eims.metc.state.mn.us/Download?startDate=01-01-0001&endDate=12-31-9999&parameterIds=2107;2108;2135;2106;2190;2198;2240;2272;2243;2244;2245;2246;2191;2196;2061;2062;2063;2064;2065;2066;2078;2079;2080;2081;2082;2083;2103;2104;2105;2174;2207;2185;2186;2194;2211;2220;2248;2273;2197;2208;2212&siteTypes=River/Stream&majorWatersheds=Blue%20Earth%20River;Cannon%20River;Le%20Sueur%20River;Lower%20Minnesota%20River;Lower%20St.%20Croix%20River;Minnesota%20River%20-%20Mankato;Mississippi%20River%20-%20Lake%20Pepin;Mississippi%20River%20-%20Twin%20Cities;North%20Fork%20Crow%20River;Rum%20River;South%20Fork%20Crow%20River&shortSiteIds=~0wnNIa-l9MLWV;EbEUElEPDxJuEqC6JdD7DOJ1COB5LTE1EvFrDIFNMAGFBKDECiDiB4FmGBFQFeFHNZDdF6FJEDCNFTKOJ3JnCtEYJ_JfJeDpJqLxFYJ9FwJ8DXJwDmFcNMCxDcJ6M8BqNKJlJsBDJ7DfCkNWNNCZJoMmEHDtMoM9DjM6DaMZE0DuM5NLNTNJM4NONnDNDMNINRN0N2BmDDNHCLNSNPNYJrCjNAHLN9NCNDNcM2GQGSN6NqDL;&format=csv){.uri}

## **Option 1: Read in data directly from web using EIMS query:**

Note - format is funky when i do it this way

```{r}
nutrients_url<-'https://eims.metc.state.mn.us/Download?startDate=01-01-0001&endDate=12-31-9999&parameterIds=2037;2038;2039;2163;2164;2221;2222;2223;2224;2225;2226;2172;2173;2231;2232;2233;2234;2235&siteTypes=River/Stream&majorWatersheds=Blue%20Earth%20River;Cannon%20River;Le%20Sueur%20River;Lower%20Minnesota%20River;Lower%20St.%20Croix%20River;Minnesota%20River%20-%20Mankato;Mississippi%20River%20-%20Lake%20Pepin;Mississippi%20River%20-%20Twin%20Cities;North%20Fork%20Crow%20River;Rum%20River;South%20Fork%20Crow%20River&shortSiteIds=~0INn9wla-;E1FTFrJdB5DxCOBKJuD7C6MmJ7J_EbN0GFJ1EDEHJwF6EqN2JfFYFHDOJqFQNnElDEJlGBEvJeJnJ9JoFmCtEPFeFJFcEUJsEYDpDjBqDmB4DdJrDaBDDXMACjNKBmNZCLKODtNYJ8NWCiDDDfFwFNJ3MoCZCkCNCxDLDIDiM8M6NcNMDNJ6DuNLDcNTMZM5DMLxM4NJN9NONILTM9GSGQE0M2NSNAHLNHNPNCNRNDNN;&format=csv' 

nutrients<-read.csv(nutrients_url)
head(nutrients)
#View(nutrients)

#write to file:
setwd(dir.chem)
write.csv(nutrients, paste(dir.chem, "/METC_nutrients.csv", sep=""), row.names=FALSE)
```

## **Option 2: Download data from portal using portal interface and then load into R and combine**

```{r}

#Show all files pre-downloaded from portal:
list.files(dir.chem)

#List all csv files that contain EIMS data:

files <- list.files(path = dir.chem, pattern = "EIMS.*\\.csv$", full.names = TRUE)
files

#Check column names of each file to make sure they're the same across files:
#and omit first 5 rows of metadata

colnames_list <- lapply(files, function(f) {
  names(read_csv(f, skip = 5, n_max = 0, show_col_types = FALSE))
})

# Name each element in the list with the file name (optional)
names(colnames_list) <- basename(files)

# View differences in column names across files:
colnames_list

#Add column for general parameter type
#(extract first word from file name)
get_first_word <- function(file_path) {
  file_name = basename(file_path)
  first_word <- str_extract(file_name, "^[^ _]+")
  return(first_word)
}

#Read in all EIMS csv files to a list:
file.list<-lapply(files, function(f){
  first_word = get_first_word(f)
  read_csv(f, skip=5) %>% 
    mutate(ParameterType=first_word)
}
)


# Combine all EIMS tables into one datafame
EIMS_combined <- file.list %>% 
  bind_rows()
#View(EIMS_combined)

```

## Look at all possible Met Council parameters

```{r}

levels(factor(EIMS_combined$PARAMETER))

#look at number of parameters by major parameter type

EIMS_combined %>% 
  group_by(ParameterType) %>% 
  dplyr::select(PARAMETER, ParameterType) %>% 
  distinct() %>% 
  count()

#Look at parameters in each major type:
View(EIMS_combined %>% 
  filter(ParameterType=="Nutrients") %>% 
  group_by(PARAMETER) %>% 
  count())

```

## Map locations of MetC sites

```{r}

names(EIMS_combined)

EIMS_sites_sf<-EIMS_combined %>% 
  dplyr::select(STATION_ID, NAME, X_COORDINATE_UTM, Y_COORDINATE_UTM) %>% 
  distinct() %>% 
#create lat/long coordinates in decimal degrees from UTM
  mutate(longitude= utm2lonlat(X_COORDINATE_UTM, Y_COORDINATE_UTM, zone=15, hemispher="N")$longitude) %>% 
  mutate(latitude= utm2lonlat(X_COORDINATE_UTM, Y_COORDINATE_UTM, zone=15, hemispher="N")$latitude) %>% 
#create spatial object
  st_as_sf(coords=c("longitude", "latitude"), crs=4326) %>%  #4326 is code for WGS84
#transform to NAD83
  st_transform(crs=4269)

#quick check of map:
plot(EIMS_sites_sf["NAME"])

```

Make a nicer map:

```{r}
#read in county boundaries for MN using tigris package
counties <-counties("Minnesota")
head(counties)
names(counties)

st_crs(counties) #check coordinate reference system 


#Create list of counties in TC Metro 
Metro.List<-c("Anoka", "Carver", "Dakota", "Hennepin", "Ramsey", "Scott", "Washington")

#Filter county shapefile to TC Metro only 
TC.Metro.counties<-counties %>% 
  filter(NAME %in% Metro.List)


#Add in NHD streams layer:
#First dissolve the boundaries between TC counties to create a single contiguous polygon for the area of interest (AOI)
TC_contig <- st_union(TC.Metro.counties)

NHD_TC<-get_nhdplus(
  AOI = TC_contig)

#order of agencies to show in plot
#put MPCA first so it doesn't cover up all the other agencies (where there is overlap)


#Plot the map

tmap_mode("plot")


#map colors from color brewer
color.palette<-c('#66c2a5','#fc8d62','#8da0cb','#e78ac3','#a6d854','#ffd92f','#e5c494','#b3b3b3')

EIMS.map<-
tm_shape(TC.Metro.counties)+
  tm_borders()+
tm_shape(NHD_TC)+
  tm_lines(col="#2c7bb6", lwd=0.5)+
tm_shape(EIMS_sites_sf)+
#sites sampled by agencies other than MPCA:
  #tm_shape(biodata_NAD83 %>% filter(!agency=="MPCA"))+
  tm_dots()
EIMS.map

```

# Combine WQP + EIMS Water Chemistry Data

For the entire state of MN

```{r}

#water chem data from Met Council
names(EIMS_combined)
View(EIMS_combined)

#water chem data for state of MN from WQP
names(WQP.chem.sites.specific)
View(WQP.chem.sites.specific)

#Select key columns from both datasets 
#Convert coordinates to common units
#Convert columns to common names
#Add missing columns as needed 
#Convert Date to common format 

EIMS.sel<-EIMS_combined %>% 
  mutate(OrganizationFormalName="MetCouncil") %>% 
  
  #create Longitude in decimal degrees  
   mutate(longitude= utm2lonlat(X_COORDINATE_UTM, Y_COORDINATE_UTM, zone=15, hemispher="N")$longitude) %>% 
  #create Latitute in decimal degrees:
  mutate(latitude=utm2lonlat(X_COORDINATE_UTM, Y_COORDINATE_UTM, zone=15, hemispher="N")$latitude) %>% 
#format Date
  mutate(Date=as.Date(as.POSIXct(START_DATE_TIME, format = "%m/%d/%Y %H:%M"))) %>% 
  #add additional blank columns to match WQP
  mutate(HydrologicCondition="NA", HydrologicEvent="NA", Analytical.Method="NA") %>% 
  #rename attributes
  rename(SiteName=DESCRIPTION) %>% 
 #select desired attributes
   dplyr::select(OrganizationFormalName, SiteName, latitude, longitude, Date, PARAMETER, RESULT, UNITS, HydrologicCondition, HydrologicEvent, Analytical.Method)

View(EIMS.sel)


WQP.sel<-WQP.chem.sites.specific %>% 
  #rename columns to match 
  rename(latitude=LatitudeMeasure, longitude=LongitudeMeasure, PARAMETER=parameter_nm, UNITS=parameter_units, SiteName=MonitoringLocationName, RESULT=ResultMeasureValue, Analytical.Method=ResultAnalyticalMethod.MethodName) %>%
  #format Date
  mutate(Date=as.Date(ActivityStartDate)) %>% 
  #select desired attributes
 dplyr::select(OrganizationFormalName, SiteName, latitude, longitude, Date, PARAMETER, RESULT, UNITS, HydrologicCondition, HydrologicEvent, Analytical.Method)

#View(WQP.sel)

#Combine two datasets:

Water.chem.all<-rbind(WQP.sel, EIMS.sel) 

nrow(Water.chem.all)
Water.chem.all<-Water.chem.all %>% 
  filter(!is.na(latitude))
nrow(Water.chem.all)
```

Check the map of all water chemistry locations

```{r}

#Write water chem to spatial object:
Water.chem.all.sf<- st_as_sf(Water.chem.all, coords=c("longitude", "latitude"), crs=4269)  #4269  is code for NAD83

#Quick and dirty check:
plot(Water.chem.all.sf %>% st_geometry())

#number of distinct sites
Water.chem.all %>% dplyr::select(latitude, longitude) %>% distinct() %>% count()

#Nicer map 
#read in USA state boundaries using tigris package
states<-states()
names(states)

#MN only 
MN_sf <- st_as_sf(states %>% filter(STUSPS=="MN")) #convert to sf object

tmap_mode("view")

names(Water.chem.all.sf)

#plot only unique locations

tm_shape(MN_sf) +
tm_borders()+
tm_shape(Water.chem.all.sf %>% dplyr::select(geometry) %>% distinct())+
 tm_dots()


```

Write all chemistry data to file

```{r}
write.csv(Water.chem.all, paste0(dir.output, '/WQP_EIMS_waterchem_combined.csv'))
```
