---
title: "Biodata_processing"
author: "Christy Dolph"
date: "2025-01-30"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
  chunk_output_type: console
---

# Overview

This script provides initial processing of biomonitoring datasets
received from Twin Cities metro water management organizations.

# Workspace Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Clean up workspace and load packages (or install if necessary)
rm(list=ls())

want <- c("readxl", "writexl", "tidyverse", "zoo", 
          "sf", "tmap", "purrr", "fs")
          
          need <- want[!(want %in% installed.packages()[,"Package"])]
if (length(need)) install.packages(need)
lapply(want, function(i) require(i, character.only=TRUE))
rm(want, need)

#Overall project directory:
root.dir <- getwd() 
root.dir


```

# Anoka Conservation District

## Read in data

List all excel files in the relevant agency folder

Identify excel sheets to be excluded, because of errors with data
collection

Read in and merge all remaining excel sheets

```{r}

# Set the data directory to a subdirectory you want to work with (e.g., Anoka County)
#Note ".." moves you one folder up
#Output Data Directory for TC data:
output.dir <- paste0(root.dir, "/Data/biodata_raw/TC_Metro/Compiled_data_clean") 
output.dir

#Raw data directory:
raw.dir<-paste0(root.dir, "/Data/biodata_raw/TC_Metro/Raw_data/Anoka_CD/data/RAW_Data")
raw.dir

spatial.dir<-paste0(root.dir, "/Data/biodata_raw/TC_Metro/Raw_data/Anoka_CD/data/GIS")
spatial.dir

# List all Excel files in all subdirectories (including newer and older .xlsx and .xls files)
excel_files <- list.files(raw.dir, pattern = "\\.xls[x]?$", full.names = TRUE, recursive = TRUE)
excel_files

```

```{r}
#Define search words to exclude sheets
EXCLUSION_STRINGS <- c(
  "low",
  "uncertain",
  "thrown",
  "not preserved",
  "technique",
  "not sampled",
  "reliable",
  "incomplete",
  "use",
  "inadequate",
  "decomposed",
  "rot",
  "poor"
)

# Create a single, case-insensitive regex pattern for efficient partial matching
EXCLUSION_PATTERN <- paste(EXCLUSION_STRINGS, collapse = "|")


# --- Helper Function for Single Sheet Processing ---
#Identifies sheets that were flagged for data collection problems
#Reads only the rows and columns that contain site, date and taxonomic data
#Deals with difficult format of original excel sheets 

process_single_sheet <- function(file, sheet, pattern) {
  
  # Read the sheet content: skip first 5 rows (that contain data entry instructions), and read to a maximum of 11 rows (to cover original rows 6 through 16)
  #only original rows 6:10 contain needed taxonomic data, but rows up to 16 may include notes about the quality of the sampling effort 
  
  dat <- read_excel(file, sheet = sheet, skip = 5, n_max = 11, col_types = "text")
  
  excluded_by_string <- FALSE
  processed_dat <- NULL
  
  if (nrow(dat) > 0) { #make sure sheet had data 
    # CHECK FOR EXCLUSION STRINGS
    all_values <- unlist(dat, use.names = FALSE)
    found_match <- any(grepl(pattern, all_values, ignore.case = TRUE))
    
    if (found_match) {
      excluded_by_string <- TRUE
      # processed_dat remains NULL (Excluded from final_data)
    } else {
      # Proceed with data processing logic for sheets NOT excluded by string
      if (nrow(dat) >= 4) {
        # Get values from row 6 (row 1 after skipping) for first 4 columns
        header_values <- dat[1, 1:4] |> unlist(use.names = FALSE)
        
        # Keep only the first 4 rows (original rows 6-9)
        processed_dat <- dat %>% slice(1:4)
        
        # Replace columns 1:4 #This is to make sure all replicates have Site ID and Date
        replaced_cols <- map2(processed_dat[, 1:4], header_values, ~ rep(.y, nrow(processed_dat)))
        processed_dat[, 1:4] <- bind_cols(replaced_cols)
      }
    }
  }
  
  # Return data AND the sheet's specific exclusion status
  return(list(
    data = processed_dat,
    excluded_string_match = excluded_by_string
  ))
}


# --- Main Function to Process All Sheets in a File ---
read_selected_rows <- function(file) {
  pattern <- get("EXCLUSION_PATTERN", envir = .GlobalEnv)
  sheet_names <- excel_sheets(file)
  
  # Process all sheets
  sheets_results <- map(sheet_names, ~ process_single_sheet(file, .x, pattern))
  names(sheets_results) <- sheet_names
  
  # 1. Extract the processed data frames
  sheets_data <- map(sheets_results, "data")
  
  # 2. Create the exclusion log for this file
  exclusion_log <- tibble(
    Source_File = basename(file),
    Sheet = sheet_names,
    Excluded_By_String = map_lgl(sheets_results, "excluded_string_match")
  )
  
  return(list(
    file_path = file,
    sheets_data = sheets_data,
    exclusion_log = exclusion_log
  ))
}


# --- Execution and Aggregation ---

# Read and process all Excel files using the functions above
all_data_structured <- map(excel_files, read_selected_rows)

# 1. AGGREGATE THE EXCLUSION LOG (detailed log showing which sheets had IDed data issues)
full_exclusion_log_df <- all_data_structured %>%
  map("exclusion_log") %>% # Extract the log tibble from each file result
  list_rbind() %>%         # Combine all log tibbles into a single data frame
  filter(Excluded_By_String == TRUE) %>% # Keep only the sheets that were excluded
  select(Source_File, Sheet)

# 2. COMBINE THE PROCESSED DATA (for the 'final_data' dataframe)
all_data_flat <- all_data_structured %>% 
  map2(excel_files, ., function(file, file_result) {
    sheets <- file_result$sheets_data
    sheet_names <- names(sheets)
    
    map2(sheets, sheet_names, function(data, sheet) {
      if (is.null(data)) return(NULL) # Sheets excluded by string or skipped (short) are NULL
      data %>%
        mutate(across(everything(), as.character),
               Source_File = basename(file),
               Directory = dirname(file),
               Sheet = sheet)
    })
  }) %>% flatten() %>% compact()

# Combine all data frames into one if they have the same structure
final_data <- bind_rows(all_data_flat, .id = "ID")

View(final_data)

# --- Check excluded sheets ---

cat("\n--- List of Sheets Excluded Due to String Match (Rows 6-16) ---\n")
if(nrow(full_exclusion_log_df) == 0) {
  print("No sheets were excluded based on the string criteria.")
} else {
  print(full_exclusion_log_df)
}

#See all the excluded sheets
full_exclusion_log_df %>% 
print(n=nrow(.))

cat("\n--- List of Excel File Names Containing At Least One Excluded Sheet ---\n")
excluded_files_list <- unique(full_exclusion_log_df$Source_File)
if(length(excluded_files_list) == 0) {
  print("No files contained sheets excluded by the string criteria.")
} else {
  print(excluded_files_list)
}
```

## Format and clean up data:

```{r}
#Additional formatting to fix tables

names(final_data)

#Identify columns you don't need:
not.needed<-c("ID", "# individuals", "Mean # individuals per replicate", "EPT", "Family biotic index", "Number of Families", "Dominant family", "Dominant family% overall", "% comp Ephemeroptera", "% comp Plecoptera", "% compTrichoptera", "% comp Diptera", "% comp Coleoptera", "% comp Oligochaeta", "% comp Other", "Source_File", "Directory",  "squares in grid", "squares sampled", "Number of Replicates")   

#note dates are in multiple formats across Excel sheets, so will have to check for those 



final_data_formatted <- final_data %>%
  # format date column
  rename(Date = "Sampling Date") %>%
  
  # *** CHECK FOR MULTIPLE DATE FORMATS ***
  mutate(
    # 1. Clean up known problematic date strings BEFORE conversion
    Date_Cleaned = str_replace(Date, "31 May 2011 and 6 June 2011", "2011-06-06"),
    
    # Now use the cleaned column for conversion
    # 2. Try to convert as Excel Serial Number
    Collection_Date_Numeric = suppressWarnings(as.Date(as.numeric(Date_Cleaned), origin = "1899-12-30")),
    
    # 3. Try to convert as Day-Month-Year (robust for "12 Oct. 2011")
    Collection_Date_DMY = suppressWarnings(lubridate::dmy(Date_Cleaned)),
    
    # 4. Try to convert as Month-Day-Year (for other text formats)
    Collection_Date_MDY = suppressWarnings(lubridate::mdy(Date_Cleaned)),
    
    # 5. Combine and prioritize: Use Numeric, then DMY, then MDY
    Collection_Date = coalesce(
      Collection_Date_Numeric,
      Collection_Date_DMY,
      Collection_Date_MDY
    )
  ) %>%
  # Remove temporary date columns
  select(-Date_Cleaned, -Collection_Date_Numeric, -Collection_Date_DMY, -Collection_Date_MDY) %>% 
  # *** END DATE LOGIC ***
  
  # remove rows where no invert IDs were collected (ie, where no replicates were collected)
  # NOTE: Be careful with the positional indices (10:95 etc); it's fragile
  filter(!if_all(c(10:97, 117:159, 162:178), is.na)) %>%
  # fill in Site, Collection_Date and other information for blank rows:
  mutate(across(c(`Site Name`, Collection_Date, `Who conducted sampling?`), zoo::na.locf)) %>%
  
  # remove unneed columns, including the "Families --->" header column
  dplyr::select(!`Families  --->`) %>% 
  dplyr::select(!all_of(not.needed)) %>%
  # change family counts to long form
  pivot_longer(
    cols = -c(`Site Name`, Date, Collection_Date, # <-- Included the new date column
             `Who conducted sampling?`, `Replicate Number`, `Squares in grid`,
             `Squares Sampled`, `Q/A performed?`, Id_Comments),
    names_to = "Family",
    values_to = "count"
  ) %>%
  rename(
    SiteID = `Site Name`,
    Data_collectors = `Who conducted sampling?`
  ) %>% 
  mutate(Agency="Anoka Conservation District")

View(final_data_formatted)

```

## Check years and sites for which data is available

```{r}
View(final_data_formatted %>% 
  group_by(SiteID) %>% 
  summarise(min.date=min(Collection_Date), max.date=max(Collection_Date)))


#Note discrepancies in Site Names! (Site Names entered slightly differently in some years). Some of the discrepancies are quite confusing (eg, two different Hardwood Creek locations on Hwy 140 vs Co Rd 140)

View(final_data_formatted %>% 
  filter(Collection_Date==as.Date("2010-10-29")))

```

## Load in spatial data (lat/longs) for sampled sites

```{r}
#load in site locations:  

locations<-st_read(file.path(spatial.dir, "ACD_Biomonitoring_All_Years.shp")) 

```

Plot site locations:

```{r}

tm_shape(locations) +    
  tm_dots()+   
  tm_text("SITEID", size=0.5)  

#Note near duplicate locations for some sites (eg, RumRiver_Hwy24)
```

Clean up and look at SiteIDs in spatial data:

```{r}
#Remove duplicate locations 
nrow(locations)  

locations.clean<-locations %>%    
  distinct(SITEID, .keep_all=TRUE)  

nrow(locations.clean) #check # of rows remaining  

```

## Resolve site name discrepancies:

Site name discrepancies are resolved below. Inconsistent SiteIDs are
replaced with SiteIDs used in GIS data from Anoka Conservation District.

Some additional notes on names that are unresolved. For now (as of
2.17.25) I am dropping these sites if we can't resolve their ID.

- Clearwater Creek at Centennial High School and Clearwater Creek -
  these do not have an associated lat/long in GIS file. There is only 1
  Clearwater site in GIS file - assume this is all the same site.

- Rum River at Rum Central - no lat/long in GIS file (is this the same
  as RumRiver_RumRParkN?

- Hardwood Creek Site IDs are very unclear - need verification from
  Anoka Conservation District. The precise locations of the following
  SiteIDs are unknown:

  - Hardwood Cr at FL AL C

  - Hardwood Cr

  - Hardwood Creek at Hwy 140 (unclear if this is the Le Roux property
    or the second Hwy 140 location)

  - Hardwood Cr at Hwy 140 ( Le Roux Property) - Hwy 140 and Le Roux
    property are 2 separate locations

```{r}
#Make site names consistent across years (using siteID from spatial data)

#look at site names: 
sort(locations.clean$SITEID)


#Replace specific Site name versions with a consistent name
correct_name <- "RumRiver_Hwy116"
wrong.names.RumSouth <- c(
  "Rum River South", 
  "Rum River South  at Anoka High School", 
  "Rum River South  at Anoka High School", 
  "Rum River South  at Anoka High School ( Rum River at Hwy 116)", 
  "Rum River South at Anoka High School", 
  "Rum River at Anoka High School", 
  "Rum River at Anoka High School_ Weber",
  "Rum River S. of Co Rd. 116 bridge",
  "Rum River at Hwy 116",
  "Rum South  at Anoka High School",
  "Rum River Hwy 116",
  "Rum R south of C R 116 bridge ( at Anoka High School)",
  "Anoka High School Rum River South",
  "Anoka High School Rum River",
  "Rum River" #note: verified in original spreadsheet that this was Rum River South
)

correct_name_RumNorth<-"RumRiver_Hwy24"
wrong.names.RumNorth<-c("Rum River North", 
                        "Rum River North Canoe Landing",
                        "Rum River North at Hwy 24 (park canoe landing)",
                        "Rum River at Co Rd 24, St. Francis",
                        "Rum River at Hwy 24",
                        "Rum River at Hwy 24 1", 
                        "Rum River at Hwy 24 (canoe landing at Rum North Co Park)")

correct_name_Rice<-"RiceCreek_Hwy65" 
wrong.names.Rice<-c("Rice Cr", "Rice Cr at Hwy 65", "Rice Cr at Hwy 65 ( Locke Park)")

correct_name_Pleasure<-"PleasureCreek_86thAve"
wrong.names.Pleasure<-c("Pleasure Cr 86th", "Pleasure Cr at 86th Ave N W", "Pleasure Cr at 86th Ave")

correct_name_Hardwood.LeRoux<-"HardwoodCreek_LaRouxProperty"
wrong.names.Hardwood.LeRoux<-c("Hardwood Cr at Hwy 140 ( Le Roux Property)",
                        "Hardwood Cr at La Roux Property")

#NOTE: These Hardwood Creek Site IDs are very unclear - need verification from Anoka Conservation District 

#correct_name_HardwoodALC<-"Hardwood Cr at Forest Lake ALC"
#wrong.names.HardwoodALC<-c("Hardwood Cr at FL AL C", "	
#Hardwood Cr")


correct_name_CoonCrEgret<-"CoonCreek_EgretBlvd"	
wrong.names.Coon.Egret<-c("Coon Cr at Erlandson Park",
"Coon Creet at Egret St", "Coon Cr at Egret Blvd in Erlandson Park")

correct_name_CoonLions<-"CoonCreek_LionsCoonCreekPark"
wrong.names.Coon.Lions<-c("Coon Cr at Hanson Blvd in Lions Park", "Coon Cr at Lions Coon Cr Park, Coon Rapids")

correct_name_CoonCrosstown<-"CoonCreek_CrosstownBlvd"
wrong.names.Coon.Crosstown<-c("Coon Cr at Crosstown Blvd", "Coon Cr at Andover High School ( Hanson Blvd)", "Coon Cr at Andover High School", "Coon Cr", "Coon Cr at Crosstown Blvd ( at Andover High School)") #Assuming Coon Cr is also at the High School because title of the excel file says "Andover H.S."

#Clean up data: 

final_data_cleaner<-final_data_formatted %>% 
  mutate(across(SiteID, \(x) str_replace(x, '@', 'at'))) %>% 
  mutate(across(SiteID, \(x) str_replace(x, 'Creek', 'Cr'))) %>% 
  mutate(across(SiteID, \(x) str_replace(x, 'HS', 'High School'))) %>% 
  mutate(across(SiteID, \(x) str_replace(x, '_', 'at'))) %>% 
  mutate(across(SiteID, \(x) str_replace(x, 'behind', 'at'))) %>%
#add space before capital letters:
  mutate(SiteID=gsub("([^ ])([A-Z])", "\\1 \\2", SiteID)) %>% 
  #add space before and after "at" if not there:
#mutate(SiteID=gsub("\\s*at\\s*", " at ", SiteID)) %>% 
  mutate(SiteID=gsub("(?<!\\S)(at)(?!\\S)(?!er)|(?<!\\s)(at)(?!er)", " at ", SiteID, perl = TRUE)) %>% 
#capitalize "Hwy" if it's not: 
  mutate(SiteID=gsub("hwy", "Hwy", SiteID, ignore.case = FALSE)) %>% 
#add space before and after Hwy of not there
mutate(SiteID=gsub("\\s*Hwy\\s*", " Hwy ", SiteID)) %>% 
#remove dashes
mutate(SiteID=gsub("-", "", SiteID)) %>% 
#unabbreviate A for Anoka:
mutate(SiteID=gsub("A ", "Anoka ", SiteID, ignore.case=FALSE)) %>% 
#replace "behind" with "at"
mutate(SiteID=gsub("Behind", "at", SiteID, ignore.case=TRUE)) %>% 
#replace "Cr." with "Cr"
mutate(SiteID=gsub("Cr. ", "Cr ", SiteID)) %>% 
#remove double spaces
mutate(SiteID=gsub("  ", " ", SiteID)) %>% 
#fix site names:

  mutate(SiteID = if_else(SiteID %in% wrong.names.RumSouth, correct_name, SiteID)) %>% 
  
  mutate(SiteID= if_else(SiteID %in% wrong.names.RumNorth, correct_name_RumNorth, SiteID)) %>% 
  
  mutate(SiteID = ifelse(SiteID %in% wrong.names.Rice, correct_name_Rice, SiteID)) %>% 
  
  mutate(SiteID = ifelse(SiteID %in% wrong.names.Pleasure, correct_name_Pleasure, SiteID)) %>% 

  mutate(SiteID = ifelse(SiteID %in% wrong.names.Hardwood.LeRoux, correct_name_Hardwood.LeRoux, SiteID)) %>% 
  
  mutate(SiteID = ifelse(SiteID %in% wrong.names.Coon.Egret, correct_name_CoonCrEgret, SiteID)) %>% 

  mutate(SiteID = ifelse(SiteID %in% wrong.names.Coon.Lions, correct_name_CoonLions, SiteID)) %>%    

  mutate(SiteID = ifelse(SiteID %in% wrong.names.Coon.Crosstown, correct_name_CoonCrosstown, SiteID)) %>% 

  mutate(SiteID = ifelse(SiteID %in% c("Clearwater Cr at Centerville City Hall","Clearwater Cr at Centennial High School", "Clearwater Cr"), "ClearwaterCreek_CentervilleCityH", SiteID)) %>% 
  
  mutate(SiteID= ifelse(SiteID == "Hardwood Cr at 165th St. N & 24th Ave N ( Carlson Property)", "HardwoodCreek_165thSt", SiteID)) %>% 
  
  mutate(SiteID = ifelse(SiteID == "Cedar Cr at Hwy 22", "CedarCreek_Hwy22", SiteID)) %>% 
  
  mutate(SiteID = ifelse(SiteID == "Cedar Cr at Hwy 9", "CedarCreek_Hwy9", SiteID)) %>% 
  
  mutate(SiteID = ifelse(SiteID == "Ford Brook at Hwy 63", "FordBrook_Hwy63", SiteID)) %>% 
  
  #mutate(SiteID = ifelse(SiteID == "Ditch 6 at Valley #Drive, Andover", )) #No apparent lat/long

  mutate(SiteID = ifelse(SiteID == "Sand Cr at Xeon St. N W", "SandCreek_LionsCoonCreekPark", SiteID)) %>% 
          
  mutate(SiteID = ifelse(SiteID == "Sand Cr at Xeon St. N W", "SeelyeBrook_Hwy7", SiteID)) %>% 
  
  mutate(SiteID = ifelse(SiteID == "Trott Brook at Hwy 5", "TrottBr_Hwy5", SiteID)) %>% 
  
  mutate(SiteID = ifelse(SiteID == "Seelye Brook at Hwy 7", "SeelyeBrook_Hwy7", SiteID))
  
   
#Check cleaner data: 
View(final_data_cleaner %>% 
  group_by(SiteID) %>% 
  summarise(min.date=min(Collection_Date), max.date=max(Collection_Date)))

```

**Merge sites to lat/long location.**

Note: this will drop sites for which no site location match was readily
apparent.

```{r}

Site.loc<-merge(locations.clean %>% rename(SiteID=SITEID),
                final_data_cleaner, by=c("SiteID"))
#View(Site.loc)


#Check sites retained
View(Site.loc %>% 
  group_by(SiteID) %>% 
  summarise(min.date=min(Collection_Date), max.date=max(Collection_Date)))

```

## Final columns needed for compiled data sheet:

Create columns that match format for all compiled data. And write final
table to file.

```{r}
names(Site.loc)

Data.For.File<-Site.loc %>% 
  st_drop_geometry() %>%  #drop shapefile format
  mutate(Original_ID=SiteID) %>% 
  mutate(ID="") %>% #blank unique ID column
  mutate(Year=format(Collection_Date, "%Y")) %>%  
  rename(Replicate="Replicate Number") %>% 
  rename(Collected_By="Data_collectors") %>% 
  mutate(Original_Taxon=Family) %>% 
  mutate(Class="") %>% 
  mutate(Subclass="") %>% 
  mutate(Order="") %>% 
  mutate(Family="") %>% 
  mutate(Genus="") %>% 
  mutate(Species="") %>% 
  rename(Number=count) %>% 
  mutate(Notes="Id_Comments") %>%
  mutate(Season="") %>% 
  dplyr::select(Original_ID, ID, Collection_Date, Season, Year, Replicate, Agency, Collected_By, Class, Subclass, Order, Family, Genus, Species, Original_Taxon, Number, Notes, Latitude, Longitude)

View(Data.For.File)



```

## Remove duplicates

Note there is a duplicate sheet for Coon Creek in 2015 - same data
listed in fall and spring, same issue with Rice Creek in 2021; delete
one set of duplicates

```{r}

View(Data.For.File %>% janitor::get_dupes() %>% filter(!is.na(Number)))

nrow(Data.For.File)

Data.For.File<-Data.For.File %>% distinct()

nrow(Data.For.File)

#2021 duplicates -  

View(Data.For.File %>% filter(Collection_Date=="2021-10-12"))

View(Data.For.File %>% filter(Collected_By=="Totino Grace HS; ID check by BK")) #The only site/date with this "Collected_By" - a duplicate of the original set - drop from data

Data.For.File<-Data.For.File %>% filter(!Collected_By=="Totino Grace HS; ID check by BK")

nrow(Data.For.File)
```

## Write clean data to file

```{r}
#Write cleaned Anoka data to file:

write_xlsx(Data.For.File, file.path(output.dir, "final_data_compiled_anoka.xlsx"))
```

# Bassett Creek Watershed District

Each Bassett Creek file contains multiple sheets with a mix of tidy and
untidy formats over the years. The following script addresses these
idiosyncracies by site (8 sites total)

Google Gemini was used to assist in code development for this section
(tho verbatim Google Gemini suggestions were not used; they were
adapted)

## Read in all excel sheets for all sheets:

```{r}

#Set directory for Bassett Creek data
bassett.dir<-paste0(root.dir, "/Data/biodata_raw/TC_Metro/Raw_data/Bassett_Creek/Data")
bassett.dir

#See all excel files
list.files(bassett.dir)

#Create list of all excel files only in this folder
bass_excel_files <- list.files(bassett.dir, pattern = "\\.xls[x]?$", full.names = TRUE, recursive = FALSE)
bass_excel_files


#Create a function to read all sheets from al files into a list

read_all_sheets <- function(file_path) {
  sheet_names <- excel_sheets(file_path)
  
  # Read each sheet into a list of data frames
  sheets_data <- map(sheet_names, ~read_excel(file_path, sheet = .x))
  
  # Optionally, name the list elements with the sheet names
  names(sheets_data) <- sheet_names
  
  return(sheets_data)
}

all_excel_data <- map(bass_excel_files, read_all_sheets)

# Optionally, name the outer list with the file names
names(all_excel_data) <- basename(bass_excel_files)
  
all_excel_data

```

Process data separately by Site

Sweeney Branch

```{r}
# Define the list of all macroinvertebrate data files for this site
invertebrate_files <- c(
    SweeneyBranch.xlsx$`2020 Invertebrate Data-MPCA`,
    "SweeneyBranch.xlsx - 2015 BCWMC Invertebrate Data.csv",
    "SweeneyBranch.xlsx - 2012.csv",
    "SweeneyBranch.xlsx - 2009.csv",
    "SweeneyBranch.xlsx - 2006.csv",
    "SweeneyBranch.xlsx - 2003.csv",
    "SweeneyBranch.xlsx - 2000.csv",
    "SweeneyBranch.xlsx - 1996.csv"
)

# --- 1. Function to process the 2020 MPCA file (Tidy Format) ---
# This file is relatively clean and has standard headers.
process_2020_data <- function(file_path) {
    read_csv(file_path) %>%
        # Select and rename required columns
        select(
            SiteID = FieldNum,
            TaxonName = Name1,
            TaxonName2 = Name2,
            Count = `Transformed Count`,
            Date = DateEntered
        ) %>%
        # Combine the two taxonomic name columns
        mutate(
            TaxonomicID = if_else(
                is.na(TaxonName2) | TaxonName2 == "",
                TaxonName,
                paste(TaxonName, TaxonName2)
            ),
            # Standardize Site ID (20UM011 corresponds to Sweeney Branch)
            SiteID = "Sweeney Branch (MPCA 2020)"
        ) %>%
        # Select final columns and convert Date to standard format
        select(SiteID, TaxonomicID, Count, Date) %>%
        mutate(Date = ymd(Date))
}

# --- 2. Function to process the 2015 BCWMC file (Tidy Format) ---
# This file is also relatively clean with different headers.
process_2015_data <- function(file_path) {
    read_csv(file_path) %>%
        # Select and rename required columns
        select(
            SiteID = Sample_Station_Name,
            TaxonomicID = Taxon,
            Count = Subsample, # Using Subsample as the primary count
            Date = Sample_Date_Collected
        ) %>%
        # Select final columns and convert Date to standard format
        select(SiteID, TaxonomicID, Count, Date) %>%
        mutate(Date = mdy(Date)) # Assuming MM/DD/YYYY or similar for 2015
}

# --- 3. Function to process Hierarchical/Untidy files (1996-2012) ---
# These files have header information embedded in the first few rows,
# and the taxonomy is listed hierarchically in the first column, with counts
# in the rightmost column.
process_hierarchical_data <- function(file_path) {
    # 1. Read header info to extract Date and SiteID
    header_lines <- readLines(file_path, n = 5, warn = FALSE)

    # SiteID is consistently "Sweeney Lake Branch at Turner's Crossroad" or similar
    site_id <- "Sweeney Branch at Turner's Crossroad"

    # Extract date - look for a date pattern in the first few lines
    date_match <- str_extract(header_lines, "\\d{4}-\\d{2}-\\d{2}|\\d{1,2} [A-Za-z]{3}\\. \\d{4}|\\d{1,2}/\\d{1,2}/\\d{2,4}")
    date_match <- date_match[!is.na(date_match)]
    
    # Use the first valid date found, or default if none found
    if (length(date_match) > 0) {
        # Clean up date formats and convert
        date_extracted <- date_match[1] %>% 
          str_replace_all("\\.", "") %>% 
          mdy_or_y(.)
    } else {
        # Extract year from filename as a fallback (e.g., "2000")
        year_extracted <- str_extract(file_path, "\\d{4}")
        date_extracted <- paste0(year_extracted, "-10-01") %>% ymd()
    }

    # 2. Read the main data body, skipping initial rows and ignoring headers
    # We skip 8 rows to start at the Taxa/Count data for most files (e.g. 2012)
    data_raw <- read_csv(
        file_path,
        skip = 8,
        col_names = FALSE,
        col_types = cols(.default = col_character()),
        show_col_types = FALSE
    )

    # 3. Clean and process the data frame
    data_clean <- data_raw %>%
        # The taxon name is always in the first column (X1)
        # The count is always in the last column that contains numeric data
        select(
            TaxonomicID = 1,
            Count = last_col() # Select the last column as the count
        ) %>%
        # Convert Count to numeric and remove non-numeric entries (headers, NAs)
        mutate(
            Count = as.numeric(Count),
            TaxonomicID = str_trim(TaxonomicID) # Remove leading/trailing spaces
        ) %>%
        # Filter for rows that have a valid count (macroinvertebrate records)
        filter(!is.na(Count) & Count > 0) %>%
        # Add the extracted SiteID and Date
        mutate(
            SiteID = site_id,
            Date = date_extracted
        ) %>%
        # Select final columns and ensure order
        select(SiteID, TaxonomicID, Count, Date)
    
    return(data_clean)
}

# --- 4. Main consolidation script ---

# Initialize a list to hold the processed data frames
all_data_list <- list()

# Loop through all files and apply the appropriate function
for (file in invertebrate_files) {
    print(paste("Processing:", file))

    if (str_detect(file, "2020")) {
        # Apply 2020-specific function
        df <- process_2020_data(file)
    } else if (str_detect(file, "2015")) {
        # Apply 2015-specific function
        df <- process_2015_data(file)
    } else {
        # Apply the complex hierarchical function (1996 - 2012)
        df <- process_hierarchical_data(file)
    }

    # Add the result to the list
    all_data_list[[file]] <- df
}

# Combine all data frames into a single, comprehensive table
compiled_invertebrate_data <- bind_rows(all_data_list)

# --- 5. Output and final clean up ---

# Export the final compiled data to a new CSV file
# This will be in the same folder as your input files.
write_csv(compiled_invertebrate_data, "compiled_macroinvertebrate_data.csv")

print("--- Data Compilation Complete! ---")
print(paste("Total Records Compiled:", nrow(compiled_invertebrate_data)))
print("Output file saved as: compiled_macroinvertebrate_data.csv")
print("First few rows of the compiled data:")
print(head(compiled_invertebrate_data))

# Also print a summary of how many records were found per year
print("Records per year:")
compiled_invertebrate_data %>%
    mutate(Year = year(Date)) %>%
    group_by(Year) %>%
    summarise(TotalRecords = n())
```

# Dakota County

## Read in all files

Columns are the same for every year

```{r}

#read in all Dakota files

#Set directory for Dakota data
dakota.dir<-paste0(root.dir, "/Data/biodata_raw/TC_Metro/Raw_data/Dakota County Soil and Water/Raw_Data")
dakota.dir

#See all excel files
list.files(dakota.dir)

#Create list of all excel files only in this folder
dakota_excel_files <- list.files(dakota.dir, pattern = "\\.xls[x]?$", full.names = TRUE, recursive = FALSE)
dakota_excel_files


#Read in all excel files

all_data <- dakota_excel_files %>% 
  set_names(basename(.)) %>% 
  map_df(read_excel, .id = "source_file")

head(all_data)

```

## Read in coordinates for site locations and merge to data

Note that coordinates were not provided for all sampling locations, in
particular the "DUP" locations, and A16-A25

```{r}

coor.dir<-paste0(root.dir, "/Data/biodata_raw/TC_Metro/Raw_data/Dakota County Soil and Water/Coordinates")

list.files(coor.dir)

dakota.loc<-read_excel(paste0(coor.dir, "/Dakota_Coordinates.xlsx"))

head(dakota.loc)
names(dakota.loc)

#Check site names
levels(factor(dakota.loc$Site))
levels(factor(all_data$Sample_Client_ID))

#Site naming discrepancies - 
#Assuming "A01" is the same as "A1" etc

#Rename site locations for consistency (removing 0s for single digit numbers)

all_data<-all_data %>%
  dplyr::rename(Site=Sample_Client_ID) %>% 
  mutate(Site=str_replace(Site, "A0(?=\\d)", "A"))

levels(factor(all_data$Site))

#Merge with coordinates
#Use downstream coordinates (end of sampling reach)

all_data_coord<-merge(all_data, dakota.loc, by=c("Site"))
nrow(all_data_coord)

nrow(all_data) #lost about 300 rows 

#Format names to match other datasets
names(all_data_coord)

View(all_data_coord)

all_data_coord<-all_data_coord %>% 
  dplyr::rename(Collection_Date=Sample_Date_Collected, Latitude=Down_Latitude, Longitude=Down_Longitude, Original_ID=Site, Number=Subsample, Original_Taxon=Taxon) %>% 
  mutate(Agency="Dakota County SWCD") %>% 
  mutate(Year=year(as.Date(Collection_Date, format="%Y-%m-%d")))

#List of target columns for procecssed data
target_cols<-c("Original_ID",	"Collection_Date", "Season",	"Year",	"Replicate",	"Agency",	"Collected_by",	"Class", "Subclass", "Order", "Family",	"Genus", "Species", "Original_Taxon",	"Number",	"Notes", "TSN",	"Site", "Latitude", "Longitude", "Northing", "Easting")




```

## Write data to File

```{r}

write_xlsx(all_data_coord, paste0(output.dir, "/final_data_compiled_dakota_4_8_26_cleaned.xlsx"))

```

# Metropolitan Council (MCES)

Only including data sampled from wadeable streams in 2022 and 2023 using
MPCA SOP, as all other MCES data is for rivers collected using alternate
methods (Hester-Dendy and Ponar Dredge)

## Read in data

```{r}


#Set directory for MCES data
mces.dir<-paste0(root.dir, "/Data/biodata_raw/TC_Metro/Raw_data/MCES/Macroinvertebrates")
mces.dir

#See all excel files
list.files(mces.dir)

#Create list of wadeable streams files

mces_files<-dir_ls(mces.dir, regexp = "Streams")
mces_files

#Read in stream excel files

mces_data <- mces_files %>% 
  set_names(basename(.)) %>% 
  map_df(read_excel, .id = "source_file")

head(mces_data)
View(mces_data)
```

## Read in coordinates and merge to data

```{r}

#Set directory 
mces.coor.dir<-paste0(root.dir, "/Data/biodata_raw/TC_Metro/Raw_data/MCES/Coordinates")

list.files(mces.coor.dir)

mces.coor<-read_excel(paste0(mces.coor.dir, "/MCES biomonitoring stream station metadata_UMNrequest.xlsx"))

head(mces.coor)
#View(mces.coor)

#Site names
levels(factor(mces.coor$Station_number))

levels(factor(mces_data$Sample_Client_ID))

#Make names match
#remove leading zeroes from Station Number
mces.coor<-mces.coor %>% 
  mutate(Site = str_remove(Station_number, "00"))

#remove space and decimal from Sample Client ID
mces_data<-mces_data %>% 
  mutate(Site=str_remove(Sample_Client_ID, " ")) %>% 
  mutate(Site=str_remove(Site, "\\."))
  
#couple of other IDs to fix manually:

mces.coor<-mces.coor %>% 
  mutate(Site=ifelse(Site=="CW0231", "CW231", Site)) %>% 
  mutate(Site=ifelse(Site=="MH17", "MHC17", Site))

levels(factor(mces.coor$Site))
levels(factor(mces_data$Site))

mces.data.coord<-merge(mces_data, mces.coor, by=c("Site"))
nrow(mces.data.coord)

nrow(mces_data) #lost about 30 rows --> these were empty rows with no data, so it's OK. 

```

## Format column headings

```{r}

names(mces.data.coord)

View(head(mces.data.coord))

mces.data.coord<-mces.data.coord %>% 
  rename(Original_ID=Site, Collection_Date=Sample_Date_Collected, Original_Taxon=Taxon, Number=Subsample) %>% 
  mutate(Year=year(as.Date(Collection_Date, format="%Y-%m-%d"))) %>% 
  mutate(Agency="Metropolitan Council")

#List of target columns for procecssed data
target_cols<-c("Original_ID",	"Collection_Date", "Season",	"Year",	"Replicate",	"Agency",	"Collected_by",	"Class", "Subclass", "Order", "Family",	"Genus", "Species", "Original_Taxon",	"Number",	"Notes", "TSN",	"Site", "Latitude", "Longitude", "Northing", "Easting")
```

## Write data to file

```{r}

write_xlsx(mces.data.coord, paste0(output.dir, "/final_data_compiled_MCES_4_8_26_cleaned.xlsx"))

```

# Riley Purgatory

Original dataset processing notes:

-There is a coordinates tab, but it lacks coordinates for all sites, so
am using the coordinates listed in the stream_macro tab. Both "middle"
and "upstream" locations are listed. I assume these coordinates
correspond to the upstream start and middle locations of the sampled
reach. i will use "middle" locations for all sites.

-Only one P6 row had coordinates, assumed this was the coordinates for
all P6 samples

-lat/long for some sites was reversed, so i fixed them

-I made a 'clean' site coordinates file, by selecting coordinates from
the stream_macroinvertebrates tab and filtering them to unique values:
"site_locations_cld_4_9_26.xlsx"

-one collection was done over 2 days: 10/10/2005-10/18/2005; i will
assign later date as the date

## Macroinvertebrates

```{r}


#Set directory 
riley.dir<-paste0(root.dir, "/Data/biodata_raw/TC_Metro/Raw_data/Riley Purgatory")

list.files(riley.dir)

#Read in biodata:

riley.bugs<-read_excel(paste0(riley.dir, "/stream macroinvertebrates and fish database_cldedited_4_9_26.xlsx"), sheet='stream_macro')


head(riley.bugs)
names(riley.bugs)

#Load site locations: 

riley.coor<-read_excel(paste0(riley.dir, "/site_locations_cld_4_9_26.xlsx"))
head(riley.coor)

#Merge biodata to lat/long

#First check if site names match
levels(factor(riley.bugs$site))
levels(factor(riley.coor$site))

riley.bugs.coor<-merge(riley.bugs, riley.coor %>% dplyr::select(site, Latitude, Longitude), by=c("site"))
nrow(riley.bugs.coor)

nrow(riley.bugs) #No rows lost in merge


#Select and format needed columns
target_cols

names(riley.bugs.coor)
View(riley.bugs.coor)


riley.bugs.coor<-riley.bugs.coor %>% 
  rename(Original_ID=site, Collection_Date=date, Original_Taxon=species) %>% 
  mutate(Agency="Riley-Purgatory")

#Write to file
write_xlsx(riley.bugs.coor, paste0(output.dir, "/final_data_compiled_riley_purgatory_4_9_26_cleaned.xlsx"))
```

## Fish data

Note: not all fish data has date collected, but it does have year

```{r}

riley.fish<-read_excel(paste0(riley.dir, "/stream macroinvertebrates and fish database_cldedited_4_9_26.xlsx"), sheet='stream_fish')


levels(factor(riley.fish$Site))

riley.fish.coor<-merge(riley.fish %>% rename(site=Site), riley.coor %>% dplyr::select(site, Latitude, Longitude), by=c("site"))
nrow(riley.fish.coor) 

nrow(riley.fish) #lost ~45 rows 

names(riley.fish.coor)

#format columns
target_cols

riley.fish.coor$Collection_Date <- make_date(year = riley.fish.coor$Year, month = riley.fish.coor$Month, day = riley.fish.coor$Day)

riley.fish.coor<-riley.fish.coor %>% 
  rename(Original_ID=site, Original_Taxon=`common name`) %>% 
  mutate(agency="Riley-Purgatory")


#Write to file
write_xlsx(riley.fish.coor, paste0(output.dir, "/Fish_data_cleaned/FISH_final_data_compiled_riley_purgatory_4_9_26_cleaned.xlsx"))
```
