---
title: "WQP_data_retrieval_script"
author: "Christy Dolph"
date: "2024-06-03"
output: html_document
editor_options: 
  chunk_output_type: inline
---

# Accessing data from the USGS Water Quality Portal (WQP)

Updated: 9.8.25

Note: dataRetrieval functions have changed substantially:

<https://doi-usgs.github.io/dataRetrieval/articles/read_waterdata_functions.html>

## Data in the new format now available:

<https://waterdata.usgs.gov/blog/wqx3/>

## Use the dataRetrieval package to access WQP

See [https://github.com/DOI-USGS/dataRetrieval](https://github.com/DOI-USGS/dataRetrieval?tab=readme-ov-file)

```{r setup, include=FALSE}
##update and install packages as needed 


knitr::opts_chunk$set(echo = TRUE)

# Clean up workspace 
rm(list=ls())

#install and load packages

#list of necessary packages 
want <- c('dataRetrieval', 'tidyverse', 'sf', 'tmap', 'tigris', 'purrr')


need <- want[!(want %in% installed.packages()[,"Package"])]
if (length(need)) install.packages(need)
lapply(want, function(i) require(i, character.only=TRUE))
rm(want, need)


#Install StreamCatTools (if not already installed):
# First install remotes if needed
if (!requireNamespace("remotes", quietly = TRUE)) {
  install.packages("remotes")
}
# Install StreamCatTools
if (!requireNamespace("StreamCatTools", quietly = TRUE)) {
  remotes::install_github("USEPA/StreamCatTools")
}
# Load the package
library(StreamCatTools)



# Create subdirectory shortcuts

#Main directory
dir.root<-getwd()

#Raw data
dir.raw<-paste0(dir.root, "/data/raw_data")

```

# Discovering data

You can specify site type, parameter/characteristic (ie, wq parameter) type, time period, and spatial scope (eg, individual states), etc. These are a bit finicky to figure out!

To find all available parameter options you can either view the documentation at: <https://www.waterqualitydata.us/webservices_documentation>

OR you can go to the web-based WQP, and use the form to discover which parameters are available. Once the query is set, scroll down to the "Query URL" to ID the parameters:

<https://www.waterqualitydata.us/#advanced=true>

## Get list of all parameters using dataRetrieval package

Trying to find IBI data!

```{r}

#Read in all possible parameter codes:

parameterData <- read_waterdata_parameter_codes()

View(parameterData)


#Check groups of parameters
levels(factor(parameterData$parameter_group_code))

#Check biology and population parameter codes
View(parameterData %>% filter(parameter_group_code=="BIO"))

#Benthic invertebrates, tot: no. individuals per square meter
#parameter code = 70943

View(parameterData %>% filter(parameter_group_code=="POP"))
#POP# is for 'population/community' i believe 

```

## Discover sites where a certain type of data exists

Note this does not provide info about whether sites were repeatedly sampled, just that some data exists at these sites.

It does provide Lat/Long

```{r}

#For Benthic Macroinvertebrates
MN.bug.sites<-whatWQPsites(statecode= "MN", siteType ="Stream",  assemblage="Benthic Macroinvertebrates", startDateLo="1975-01-01", startDateHi="2025-01-01")

head(MN.bug.sites)
nrow(MN.bug.sites) #173 records

#This is far fewer sites than we have received from MPCA data requests!
#Evidence that not all available data is in the WQP

#Check agencies that submitted this data 
names(MN.bug.sites)
levels(factor(MN.bug.sites$OrganizationFormalName)) #interesting!

#Note: MPCA data not located here!

#How about fish:
MN.fish.sites<-whatWQPsites(statecode= "MN", siteType = "Stream", assemblage = "Fish/Nekton", startDateLo="1975-01-01", startDateHi="2025-01-01")
nrow(MN.fish.sites)

#Check agencies:
levels(factor(MN.fish.sites$OrganizationFormalName))

```

### Could macroinvert data be anywhere else?

(ie, what other search criteria might it be under?)

```{r}

#Check in assemblage "invertebrates"

MN.inverts<-whatWQPsites(statecode= "MN", siteType ="Stream",  assemblage="Invertebrates", startDateLo="1975-01-01", startDateHi="2024-01-01")

nrow(MN.inverts) #159 sites

#Check agencies 
names(MN.inverts)
levels(factor(MN.inverts$OrganizationFormalName))
#includes data from EPA NARS, Grand Portage Reservation, and Red Lake DNR

```

Other places I looked for macroinvert data:

"TaxonomicName" - does contain some info on Chironomids

# Create maps of data available from different sources/search terms

```{r}

#combine sources of macroinvert data
invert.sites<-rbind(MN.bug.sites, MN.inverts)

#create spatial data for macroinvert sampling locations from lat/longs
#Converting to `sf` Object
names(invert.sites)
invert.sites_sf <- st_as_sf(invert.sites, coords = c("LongitudeMeasure","LatitudeMeasure"))
invert.sites_sf

#Set coordinate reference system to wgs84
st_crs(invert.sites_sf) = 4326 #set crs to wgs84
st_crs(invert.sites_sf)

st_bbox(invert.sites_sf) #bounding box
#can ID spatial extent of dataset
```

### Read in state borders:

```{r}

#read in US state boundaries using tigris package
#For area of interest, select states that overlap with measured P data
#read in USA state boundaries using tigris package
states<-tigris::states()
states_sf <- st_as_sf(states) #convert to sf object
st_bbox(states_sf) #bounding box
states_sf

#reproject to WGS84 to match BCG layer
USA_WGS84 <- st_transform(states_sf,
                                st_crs(invert.sites_sf))
st_crs(USA_WGS84) #check projection 
```

### Plot locations of WQP invert data

helpful cheatsheet for palette colors: <https://stackoverflow.com/questions/57153428/r-plot-color-combinations-that-are-colorblind-accessible>

```{r}
names(USA_WGS84)
MN_sf<- USA_WGS84 %>% filter(NAME=="Minnesota")

names(invert.sites_sf)

#make nicer labels for agencies
levels(factor(invert.sites_sf$OrganizationFormalName))

invert.sites_sf$OrganizationFormalName<-factor(invert.sites_sf$OrganizationFormalName, levels=c("EPA National Aquatic Resources Survey (NARS)", "EMAP-Great Rivers Ecosystems","National Park Service Water Resources Division", "Grand Portage Reservation" , "Red Lake DNR", "Upper Sioux Community" , "Wisconsin Department of Natural Resources", "21GPBCH - Grand Portage Beaches"), labels=c("EPA National Aquatic Resource Surveys", "EPA EMAP - Great Rivers Ecosystems", "National Park Service", "Grand Portage Reservation", "Red Lake DNR", "Upper Sioux Community", "Wisconsin DNR", "Grand Portage Beaches"))

WQP.invert.map<-
  tm_shape(MN_sf)+
   tm_borders()+
 tm_shape(invert.sites_sf)+
  tm_dots(col="OrganizationFormalName",  title = "", 
          palette = palette.colors(palette = "R4"))+
   tm_layout(legend.outside = TRUE, legend.outside.size = 0.7, title="Agency")
print(WQP.invert.map) 

# Save the map as an image file
tmap_save(tm = WQP.invert.map, filename = "MN_WQP_invert_sites_map.jpg", dpi=300, width=6, height=4)

```

## Compare WQP locations to locations from previous data requests to states (including MPCA)

```{r}
library(readxl)
#Load in pre processed BCG data (for inverts)
setwd("C:/Users/dolph/OneDrive/Documents/EPA_Project_2/BCG-analysis/Preliminary_analysis_output")

BCG<-read_xlsx("./bcg_data_filtered.xlsx")
#head(BCG)

#create spatial data for BCG locations from lat/longs
#Converting to `sf` Object

#Assign "BCG" to either fish or inverts

BCG_sf <- st_as_sf(BCG, coords = c("Long","Lat"))
BCG_sf

#Set coordinate reference system to wgs84
st_crs(BCG_sf) = 4326 #set crs to wgs84
st_crs(BCG_sf)

st_bbox(BCG_sf) #bounding box
#can ID spatial extent of dataset

```

### Map of invert BCG locations in MN

```{r}
#Subset MN data
MN_BCG<-subset(BCG_sf, State=="MN")
nrow(MN_BCG) #4192 sites (note this could include repeat visits)

names(MN_BCG)
MN_BCG %>% select(SiteID) %>% distinct() %>% count() #3319 unique sites

MPCA.EPAprojectrequest.map<-
  tm_shape(MN_sf)+
   tm_borders()+
 tm_shape(MN_BCG)+
  tm_dots()
#+tm_layout(legend.outside = TRUE, legend.outside.size = 0.7, title="Agency")
print(MPCA.EPAprojectrequest.map) 

# Save the map as an image file
tmap_save(tm = MPCA.EPAprojectrequest.map, filename = "C:/Users/dolph/OneDrive/Documents/Proposals/LCCMR_2024/MN_EPAprojectrequest_invert_sites_map.jpg", dpi=300, width=6, height=4)

tmap_arrange(WQP.invert.map, MPCA.EPAprojectrequest.map)

```

Combine MPCA and WQP data in one map (for LCCMR research addendum)

```{r}
names(MN_BCG)
names(invert.sites_sf)

#Combine the two datasets for mapping

MN_BCG_simp<-MN_BCG %>% select(geometry) %>% 
  mutate(OrganizationFormalName="Minnesota Pollution Control Agency")
head(MN_BCG_simp)
WQP_invert_simp<-invert.sites_sf %>% select(OrganizationFormalName, geometry)

Bugs.all<-rbind(MN_BCG_simp, WQP_invert_simp)

palette1<-c('#b35806','#e08214','#fdb863','#fee0b6','black','#d8daeb','#b2abd2','#8073ac','#542788')

All.invert.map<-
  tm_shape(MN_sf)+
   tm_borders()+
  tm_shape(Bugs.all)+
  tm_dots(col="OrganizationFormalName",  title = "", size=0.05, palette = palette1, legend.show=FALSE)+
   tm_add_legend(type = "symbol", 
                  col = palette1,
                  labels = c(levels(factor(Bugs.all$OrganizationFormalName))))+
  tm_layout(legend.outside = TRUE, legend.outside.size = 0.6, title="Agency", legend.text.size=1.5, title.size=1.5)
print(All.invert.map) 

c(levels(factor(Bugs.all$OrganizationFormalName)))

# Save the map as an image file
tmap_save(tm = All.invert.map, filename = "C:/Users/dolph/OneDrive/Documents/Proposals/LCCMR_2024/MN_MPCA_and_WQP_invert_sites_map.jpg", dpi=300, width=6, height=4)
```

## Compare to data in MN from Twardochleb et al 2020

Freshwater insect database for CONUS

EDI data repository: <https://portal.edirepository.org/nis/mapbrowse?packageid=edi.481.5>

Github: <https://github.com/aquaXterra/freshwater_insects_CONUS>

Paper: <https://onlinelibrary.wiley.com/doi/full/10.1111/geb.13257> 

Accessed 6.4.24

```{r}
#read in raw community data from Twardochelb et al

setwd('C:/Users/dolph/OneDrive/Documents/EPA_Project_2/BCG-analysis/Taxonomic_data_raw')

Database<-read.csv('Raw_Community_Data_Twardochleb.csv', sep=",", header=TRUE)

#Look only at site locations in MN for now
names(Database)
head(Database2)
DatabaseMN<-Database %>% 
  select(Location_ID, Date, Latitude, Longitude, Data_source, Monitoring_organization, Study_state, Sample_method) %>% 
  distinct() %>% 
  filter(Study_state=="Minnesota")
nrow(DatabaseMN)

```

### Create map of Twardochelb sample locations in MN

```{r}
#Assign "BCG" to either fish or inverts

DatabaseMN_sf <- st_as_sf(DatabaseMN, coords = c("Longitude","Latitude"))
DatabaseMN_sf

#Set coordinate reference system to wgs84
st_crs(DatabaseMN_sf)
st_crs(DatabaseMN_sf) = 4326 #set crs to wgs84

#Map Twardochleb database sample locations in MN
#Overlay with sites from our own MPCA data request

Database.map<-
  tm_shape(MN_sf)+
   tm_borders()+
  tm_shape(MN_BCG)+
  tm_dots(col="green")+
  tm_shape(DatabaseMN_sf)+
  tm_dots()
#+tm_layout(legend.outside = TRUE, legend.outside.size = 0.7, title="Agency")
print(Database.map) 

#Note that Twardochleb appears to only contain data from Mississippi River Basin, as opposed to whole state 

#Nearly all sites included in Twardochleb et al database are included in our MPCA data request from 2016 (note that there may be newer MPCA sampling locations now, from 2016-2024)

```

# Pull stream taxonomic and IBI data from WQP

Overall search tearms for the readWQPdata() function were ferreted out using the search terms from the WQP advanced search web portal: <https://www.waterqualitydata.us/#advanced=true>

**Taxonomic (count) data**: querying for taxnomic data for a large area (ie, specifying state code but not county code) seems to fail by returning a bad request. Assuming this data request is too large. However, requests by specific county are successful. So, the code below create a function to query data by county in Minnesota, and then maps (loops) that code through every county. This takes several minutes.

**IBI/metric data?** (if anywhere)

```{r}

# Query WQP for biological results  
#For one county: (takes only a few seconds) #leaving assemblage unspecified gives invertebrate and fish data  

county.data <- readWQPdata(   
  statecode   = "US:27", # Minnesota   
  countycode  = "003",   # Anoka County   
  #startDateLo = "1975-01-01", #can leave date empty to query all dates   
  #startDateHi = "2025-01-01",    
  siteType="Stream", #Stream data only
  dataProfile = "biological", #"biological" includes additional meta data (such as sampling method, lat/longs), "narrowResult" gives a more simplified table
  characteristicName="Count", #for taxonomic count data
  providers   = c("NWIS","STORET") )    

View(county.data)
#For larger data query (a whole state), loop through counties one by one and then combine 
#If you lave 'countycode' empty the query crashes   

#First create list of FIPS codes for MN counties (27 is MN state code, county is 3-digit string)  

mn_counties <- sprintf("%03d", 1:87)  # 87 counties 
mn_counties   

#Function to query WQP:  

get_county_data <- function(county_fips) {   
  message("Downloading data for county ", county_fips, "...")   
  tryCatch({     
  
  df <- readWQPdata(       
  statecode   = "US:27",      # Minnesota
  countycode  = county_fips,  # County FIPS       
  siteType="Stream",
  dataProfile = "biological",       
  characteristicName= "Count",       
  providers   = c("NWIS","STORET"))          

    #Convert all columns to character so bind_rows won't choke if column types don't match across counties     
    df %>% mutate(across(everything(), as.character))        }, 
  
  error = function(e) {     
    message("  Failed for county ", county_fips, ": ", conditionMessage(e))     
    return(NULL)   }) 
  }


# Apply function to each county and then combine data #All columns are converted to character so there aren't class mismatches 
#map_dfr will read in all columns in every county dataset, and if there is no data in a particular county dataset, will add 'NA' #this is needed because columns don't all match across county datasets  

#Track how long it takes 
Start.time<-Sys.time() Start.time  

#Apply WQP data request function to all counties 
mn_data <- mn_counties %>%   
  map_dfr(get_county_data)   # fills NAs automatically  

End.time<-Sys.time() 
End.time  
End.time-Start.time #14 minutes   

#Look at data: 

View(mn_data)  

#Write to file so you can examine more easily
write.csv(mn_data, paste0(dir.raw, '/WQP_taxonomic_count_MN.csv'))  


```

Subset attributes:

```{r}

#read in assemblage data saved to file: 
mn_data<-read.csv(paste0(dir.raw, '/WQP_taxonomic_count_MN.csv'))
head(mn_data)

#Select most important columns names(mn_data)  
#Note there might be some habitat data in here   

#Subset attributes:   

mn_data_subset<-mn_data %>%    
  dplyr::select(OrganizationIdentifier, OrganizationFormalName, ActivityStartDate, ActivityLocation.LatitudeMeasure, ActivityLocation.LongitudeMeasure, AssemblageSampledName, SubjectTaxonomicName, ResultMeasureValue, CharacteristicName, ResultMeasure.MeasureUnitCode, ActivityTypeCode, ActivityMediaName, SampleCollectionMethod.MethodIdentifier, SampleCollectionMethod.MethodName, SampleCollectionMethod.MethodDescriptionText, SampleCollectionEquipmentName, SampleCollectionMethod.MethodDescriptionText, NetTypeName, BiologicalIntentName, ResultAnalyticalMethod.MethodIdentifier, ResultAnalyticalMethod.MethodName, LaboratoryName, HabitName, TaxonomicPollutionTolerance, TaxonomicPollutionToleranceScaleText, TrophicLevelName, FunctionalFeedingGroupName)  

View(mn_data_subset)
```

## Explore taxonomic data - collection methods

Plan: keep data if collection methods are similar to MPCA (most dominant collection method in state): Qualitative Multi Habitat Sampling with Dipnet/Kicknet for benthic macroinvertebrates, and Electroshocking for fish.

Otherwise omit. Additional details for data collected by specific methods below.

Also omit lake data (keeping stream data only). Note that a WQP query with siteType = "Stream" still seems to return lake data

Need to be careful about abundance estimates from these samples; relative abundance might be OK.

Note: lake data still coming up, even though streams was specified in WQP query

Searching data by assemblage and method. Notes below are for SampleCollectionMethod.MethodName

**DATA TO KEEP:**

-   "Modified D-frame kick net". EPA NARS 2004 data. Sampling based on EPA RBP protocol.

-   "NRSA Kick net sample". EPA NARS 2019 data.

-   "Fish collection by electrofishing" - EMAP Great Rivers Survey. Boat mounted-electroshock

-   "NCA Standard Fish Sample" - NARS 2008-2009 and NARS 2013-2014 fish assessments, back pack & boat mounted electroshock (for wadeable vs non-wadeable)

-   "Standard Sampling Methods" - Red Lake DNR fish samples - boat and backpack electroshock

-   "NA" (no method listed) - Fish and invertebrate data collected by Fond du Lac band and Grand Portage reservation - no methods listed? Recent samples 2024 and 2021. It might be this project? Note that this does explicitly mention stream sampling <https://forestadaptation.org/adapt/demonstration-projects/1854-treaty-authority-stream-monitoring-project>. Assuming that these are in fact stream samples (and not wetland - see note below). Would be best to follow up with 1854 Treaty Authority.

    But note: It sounds like invertebrates were also collected from Grand Portage reservation in 2021 were from wetlands (p. 18): <https://www.epa.gov/sites/default/files/2021-06/documents/grand_portage_wetland_program_plan_approved_040521.pdf> This was also a project with1854 Treaty Authority: "In 2021 and 2022, the collaborative project with the 1854 Treaty Authority will expand to include invertebrate and vegetation monitoring at regional wetlands. The methods used will be consistent with the core indicators selected for Grand Portage wetlands." This notes method for invertebrates for future studies: Genus/species level identification of light trap and pitfall trap samples." So may not be appropriate to combine with stream data. 2020 data collection notes: "2 dipnet samples (in wetlands with open water". Methods outlined in Grand Portage Band of Chippewa Wetland Assessment Project Quality Assurance Project Plan (U.S. EPA Reference Number BG 96585512, approved 5/22/20. <https://www.epa.gov/sites/default/files/2021-06/documents/grand_portage_wetland_program_plan_approved_040521.pdf>). - wetland samples likely not comparable to a stream qualitative multi habitat sample.

-   SampleCollectionMethod.MethodName=="Qualitative Multihabitat" - Red Lake DNR Qualitative Multihabitat sample w/dipnet - Inverts

**DATA TO OMIT** (tho note this data could be useful in you expand scope to lakes/wetlands)

-   SampleCollectionMethod.MethodName=="LG406". Great Lakes deepwater benthic habitats. <https://www.epa.gov/sites/default/files/2017-01/documents/sop-for-benthic-invertebrate-field-sampling-201603-8pp.pdf>

-   SampleCollectionMethod.MethodName=="NLA Standard Invert Sample". National Lake Assessment data. Kick net. (Need to check method - see documentation)

-   SampleCollectionMethod.MethodName=="Unknown". Lakes sediment benthic habitats.

-   SampleCollectionMethod.MethodName == "NLA Standard Invert Sample" - National Lakes assessment

-   SampleCollectionMethod.MethodName=="Quantitative Riffle Sample" #Red Lake DNR, D net, riffles only, 1 site only, 2016-2017, spring and fall of each year

-   SampleCollectionMethod.MethodName=="Benthic macroinvertebrate kick net sample in current")) #EMAP kick net sample in current - total \# of inverts, no taxonomic info

-   SampleCollectionMethod.MethodName=="Benthic macroinvertebrate snag net sample in current")) #EMAP snag sample - total \# of inverts, no taxonomic info

-   SampleCollectionMethod.MethodName=="USGS")) #USGS Minnesota Water Science Center - These are tissue samples (and 3 water samples)

```{r}

#Check assemblage types

mn_data_subset %>% 
  group_by(AssemblageSampledName) %>% 
  count()

#Keeping all assemblages for now 

#NA may be habitat assessment?

#Check Activity type
View(mn_data_subset %>% 
  group_by(ActivityTypeCode) %>% 
  count())


#Omit habitat assessments:
nrow(mn_data_subset)
mn_data_taxa<-mn_data_subset %>% 
  filter(!str_detect(ActivityTypeCode, "Habitat"))
nrow(mn_data_taxa)

#Check on collection methods, by assemblage
#Can data be combined (ie, are methods similar enough for data to be analyzed together?)

View(mn_data_taxa %>% 
  group_by(AssemblageSampledName, SampleCollectionMethod.MethodName) %>% 
  count())

#Look at methods for specific collections



#KEEP#########################
#Make list of datasets to keep

methods.to.keep<-c("Modified D-frame kick net", "NRSA Kick net sample", "Fish collection by electrofishing", "NCA Standard Fish Sample", "Qualitative Multihabitat")
is.na(SampleCollectionMethod.MethodName)

###########
#Exploring methods in detail:
###########

mn_data_taxa %>% 
  filter(SampleCollectionMethod.MethodName=="Modified D-frame kick net")%>% 
  summarise(minDate=min(as.Date(ActivityStartDate)), maxDate=max(as.Date(ActivityStartDate)))
  
  dplyr::select(ActivityLocation.LatitudeMeasure, ActivityLocation.LongitudeMeasure) %>% 
  distinct() %>% 
  count() #NARS 2004 survey

###
mn_data_taxa %>% 
       filter(SampleCollectionMethod.MethodName=="NRSA Kick net sample")%>% 
  summarise(minDate=min(as.Date(ActivityStartDate)), maxDate=max(as.Date(ActivityStartDate)))
  
  dplyr::select(ActivityLocation.LatitudeMeasure, ActivityLocation.LongitudeMeasure) %>% 
  distinct() %>% 
  count() #NRSA 2018-2019 survey

###
mn_data_taxa %>% 
       filter(SampleCollectionMethod.MethodName=="Fish collection by electrofishing")%>% 
  summarise(minDate=min(as.Date(ActivityStartDate)), maxDate=max(as.Date(ActivityStartDate)))
  
       dplyr::select(ActivityLocation.LatitudeMeasure, ActivityLocation.LongitudeMeasure) %>% distinct() %>% count() #EMAP Great River Ecosystem, boat electroshock

###
mn_data_taxa %>% 
       filter(SampleCollectionMethod.MethodName=="NCA Standard Fish Sample") %>% 
mutate(Year=as.Date(ActivityStartDate, format="%Y")) %>% 
  group_by(AssemblageSampledName, Year) %>% dplyr::select(ActivityLocation.LatitudeMeasure, ActivityLocation.LongitudeMeasure) %>% distinct() %>% count()
  
  summarise(minDate=min(as.Date(ActivityStartDate)), maxDate=max(as.Date(ActivityStartDate)))       
  
  #NARS 2008 and 2014 assessment, back pack and boat electroshock

###
mn_data_taxa %>% 
       filter(is.na(SampleCollectionMethod.MethodName)) %>% 
  group_by(OrganizationFormalName) %>% dplyr::select(OrganizationFormalName, ActivityLocation.LatitudeMeasure, ActivityLocation.LongitudeMeasure) %>% distinct() %>% count()
           #Fish and invertebrate data collected by Fond du Lac band and Grand Portage reservation

###
mn_data_taxa %>% 
       filter(SampleCollectionMethod.MethodName=="Qualitative Multihabitat") %>% group_by(OrganizationFormalName, AssemblageSampledName) %>% select(OrganizationFormalName, AssemblageSampledName, ActivityLocation.LatitudeMeasure, ActivityLocation.LongitudeMeasure) %>% distinct() %>% count()
  
summarise(start=min(as.Date(ActivityStartDate)), end=max(as.Date(ActivityStartDate))) 
#Red Lake DNR Qualitative Multihabitat sample w/dipnet

###
mn_data_taxa %>% 
       filter(SampleCollectionMethod.MethodName=="Quantitative Riffle Sample")
#Red Lake DNR, D net, riffles only, 1 site only, 2016-2017, spring and fall of each year

###
mn_data_taxa %>% 
       filter(SampleCollectionMethod.MethodName=="Standard Sampling Methods") %>%
       group_by(OrganizationFormalName) %>% summarise(start=min(as.Date(ActivityStartDate)), end=max(as.Date(ActivityStartDate))) 
     #  dplyr::select(OrganizationFormalName, AssemblageSampledName, ActivityLocation.LatitudeMeasure, ActivityLocation.LongitudeMeasure) %>% distinct()) #Red Lake fish - back and boat electrofihing 
  
#OMIT:

#SampleCollectionMethod.MethodName=="LG406")) #EPA Great Lakes lab, deepwater benthic sediment

#SampleCollectionMethod.MethodName=="NLA Standard Invert Sample")) #National Lakes assessment 

#SampleCollectionMethod.MethodName=="Unknown")) #EPA Great Lakes Lab, Benthic sediment

#SampleCollectionMethod.MethodName=="WATER GRAB SAMPLE")) #Wisconsin DNR, sample method unclear 

#NLA Standard Invert Sample - National Lakes assessment

#SampleCollectionMethod.MethodName=="Benthic macroinvertebrate kick net sample in current")) #EMAP kick net sample in current - total # of inverts, no taxonomic info
  
#SampleCollectionMethod.MethodName=="Benthic macroinvertebrate snag net sample in current")) #EMAP snag sample - total # of inverts, no taxonomic info

#SampleCollectionMethod.MethodName=="USGS"))
#USGS Minnesota Water Science Center - These are tissue samples (and 3 water samples)  

#UNTESTED
  
mn_data_taxa %>% 
       filter(SampleCollectionMethod.MethodName=="Standard Sampling Methods") %>%
  group_by(SampleCollectionEquipmentName) %>% dplyr::select(SampleCollectionEquipmentName) %>% distinct() #Red Lake DNR fish samples - boat and backpack electroshock

```

# Make a clean dataset of WQP taxonomic data to keep:

```{r}

methods.to.keep<-c("Modified D-frame kick net", "NRSA Kick net sample", "Fish collection by electrofishing", "NCA Standard Fish Sample", "Qualitative Multihabitat", "Quantitative Riffle Sample", "Standard Sampling Methods")
#is.na(SampleCollectionMethod.MethodName)


mn.keep<-mn_data_taxa %>% filter(SampleCollectionMethod.MethodName %in% methods.to.keep|is.na(SampleCollectionMethod.MethodName))

mn.keep %>% 
  group_by(OrganizationFormalName) %>% 
  dplyr::select(OrganizationFormalName) %>% 
  distinct()

View(mn.keep)
names(mn.keep)
#Clean up column names and identifiers

mn.keep<-mn.keep %>% 
  
  #rename columns:
  rename(Latitude=ActivityLocation.LatitudeMeasure, Longitude=ActivityLocation.LongitudeMeasure, Assemblage=AssemblageSampledName, Taxa=SubjectTaxonomicName, Date=ActivityStartDate, Count=ResultMeasureValue, Method_Orig=SampleCollectionMethod.MethodName, Method.Description=SampleCollectionMethod.MethodDescriptionText, Equipment=SampleCollectionEquipmentName, PollutionTolRef=TaxonomicPollutionToleranceScaleText) %>% 
  
  #select columns:
  dplyr::select(OrganizationFormalName, Latitude, Longitude, Date, Assemblage, Taxa, Count, Method_Orig, Method.Description, Equipment, HabitName, TaxonomicPollutionTolerance, PollutionTolRef, TrophicLevelName, FunctionalFeedingGroupName) %>%  
  
  #reclassify methods column for consistency across datasets
  mutate(Method=ifelse(Method_Orig=="Modified D-frame kick net", Method="Qualitative multi habitat, D-frame kick net", NA)) %>% 
  mutate(Method=ifelse(Method_Orig=="NRSA Kick net sample", Method="Qualitative multi habitat, D-frame kick net", Method)) %>% 
  mutate(Method=ifelse(Method_Orig=="Fish collection by electrofishing", Method="Qualitative multi habitat, D-frame kick net", Method)) %>% 
  mutate(Method=ifelse(Method_Orig=="Qualitative Multihabitat", Method="Qualitative multi habitat, D-frame kick net", Method)) %>% 
  mutate(Method=ifelse(Method_Orig=="NCA Standard Fish Sample", Method="Boat-mounted electoshock; Backpack electroshock", Method)) %>% 
  mutate(Method=ifelse(Method_Orig=="Quantitative Riffle Sample", Method="Quantitative riffle, D-net", Method)) %>% 
  mutate(Method=ifelse(Method_Orig=="Standard Sampling Methods", Method="Boat-mounted electoshock; Backpack electroshock", Method)) 
  

#Exclude NARS data as you are downloading this separately directly from EPA NARS:
##LEFT OFF 9.15.25

  

View(mn.keep)
```

Quick map:

```{r}

sites.only<-mn.keep %>% dplyr::select(ActivityLocation.LatitudeMeasure, ActivityLocation.LongitudeMeasure, OrganizationFormalName, AssemblageSampledName) %>% distinct()

sites_sf<-st_as_sf(sites.only, coords=c('ActivityLocation.LongitudeMeasure', 'ActivityLocation.LatitudeMeasure'))

sites_sf$OrganizationFormalName<-sit

tm_shape(sites_sf)+
  tm_dots(col=factor(OrganizationFormalName))

```
